---
title: "causation"
output: 
  html_document:
      toc: true
      toc_float: true
      toc_depth: 3
      number_sections: true
date: "2024-10-28"
---

# Setup

## load packages

```{r, setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

library(pacman)
pacman::p_load(tidyverse,
               ggforce,
               conflicted,
               here,
               cowplot,
               lme4,
               lmerTest,
               effects,
               Hmisc,
               performance,
               purrr,
               readr,
               igraph,
               ggraph,
               tidygraph,
               visNetwork,
               scales,
               patchwork,
               corrplot)

library(plotly) #for some reason this never loads via pacman
walk(c("select", "filter", "rename", "mutate", "summarise", "summarize"), ~ conflict_prefer(.x, "dplyr"))
conflicts_prefer(plotly::layout)
theme_set(theme_bw())
sessionInfo()

```

## read data and set variables

```{r}
desired_order <- c("see something", "hear something", "choose what to do", #please note this shows 'experience pain!' rather than feel pain
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

freesort_columns <- c("subject_id", "x", "y", "item")
# causation_columns <- c("subject_id", "itemA", "itemB", "trial_index", "original_trial_index", "response")
causation_columns <- c("subject_id", "itemA", "itemB", "response")


# data paths
directory_path <- here("code", "analysis", "pilot", "data-combined", "pilot05_121324_n50_combined_17") #data path
rdm_outputs_path <- here("code", "analysis", "pilot")
input_csv <- "data_pilot_12_14_combined"
input_path <- paste0(directory_path, "/outputs/", input_csv, ".csv") 
freesort_input_path <- paste0(directory_path, "/outputs/data_freeesort.csv")

# read data
directory_path <- here(input_path)

d <- read_csv(directory_path) %>% 
  rename("subject_id" = filename) %>% 
  mutate(
    subject_id = factor(subject_id), # Convert to factor to establish levels
    subject_index = as.integer(subject_id), # Assign a numeric index based on factor levels
    subject_id = sprintf("subj_%02d", subject_index) # Create the new sequential ID
  ) %>%
  select(-subject_index) # Remove the temporary index column if no longer needed

```

# check N

```{r}
d %>% 
  distinct(subject_id) %>% 
  count()

```


# Freesort: Plotting empirical euclidean distances

## reshape data

```{r}
#get item pairs per person
df_distances <- d %>%
  filter(trial_type == "free_sort") %>%
  select(all_of(freesort_columns)) %>%
  select(subject_id, item) %>%
  group_by(subject_id) %>%
  summarize(pairs = list(expand.grid(itemA = item, itemB = item, stringsAsFactors = FALSE))) %>%
  unnest(pairs) %>%
  filter(itemA != itemB) %>% #remove self-pairs
  left_join(d %>% 
              filter(trial_type == "free_sort") %>% 
              select(all_of(freesort_columns)), by = c("subject_id", "itemA" = "item")) %>% #append itemA positions
  rename(x_itemA = x, y_itemA = y) %>% 
  left_join(d %>% 
              filter(trial_type == "free_sort") %>% 
              select(all_of(freesort_columns)), by = c("subject_id", "itemB" = "item")) %>% #append itemB positions
  rename(x_itemB = x, y_itemB = y) %>% 
  group_by(subject_id) %>% 
  mutate(euclidean_distance = sqrt((x_itemA - x_itemB)^2 + (y_itemA - y_itemB)^2),
         emp_euclidean = ((euclidean_distance - min(euclidean_distance)) /
                                   (max(euclidean_distance) - min(euclidean_distance)))) %>% #normalizing euclidean distances
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  ungroup() %>%
  mutate(itemA = factor(itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order))

```

## plot subject-specific raw responses

```{r, fig.width = 14, fig.height = 10}
d %>% 
  filter(trial_type == "free_sort") %>%
  mutate(y = 500 - y) %>% #invert y axis
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_text(aes(label = item), size = 3, vjust = -.2) +
  facet_wrap(~subject_id)

```

## plot average raw responses

```{r}
# pdf("my_plot_average.pdf", width = 7, height = 4)  # Width and height are in inches
d %>% 
  filter(trial_type == "free_sort") %>%
  mutate(y = 500 - y) %>% 
  group_by(item) %>% 
  mutate(mean_x = mean(x),
            mean_y = mean(y)) %>% 
  ungroup() %>% 
  distinct(item, .keep_all = TRUE) %>% 
  select(-subject_id) %>% 
  ggplot(aes(x = mean_x, y = mean_y)) +
  geom_point() +
  geom_text(aes(label = item), vjust = -1) +
  labs(title = "Average Raw Responses")

```


## plot subject-specific euclidean rdms

```{r}
plot_rdm <- function(data, subject, desired_order) { # Function to create RDM plot for a single subject
  df_matrix <- data %>%
    filter(subject_id == subject) %>%
    select(-subject_id) %>%
    select(itemA, itemB, emp_euclidean) %>% 
    pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
    column_to_rownames("itemA")
  
  df_matrix <- df_matrix[desired_order, desired_order] #reorder matrix
  diag(df_matrix) <- 0  # Set diagonal to 0
  
  corrplot(as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = paste("RDM - ", subject),
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
}

subject_ids <- unique(df_distances$subject_id)

#generate plots
# plots <- map(subject_ids, ~plot_rdm(df_distances, .x, desired_order)) #run this to generate plots

# save plots with subject name as file name
 
# Generate unique subject IDs and save plots
new_names <- paste0("subj_", str_pad(seq_along(subject_ids), width = 2, pad = "0"))  # Generate new subject names

for (i in seq_along(subject_ids)) {  # Loop to save each plot as a separate PDF
  # Specify the path to save the PDF in the "figures" subfolder
  pdf(file = paste0(here(rdm_outputs_path), "/figures/freesort/", new_names[i], ".pdf"), width = 8, height = 6)  # Open a PDF device
  plot_rdm(df_distances, subject_ids[i], desired_order)  # Generate the plot
  dev.off()  # Close the device to save the file
}

```

## plot average euclidean rdm

```{r}
df_mean_distances <- df_distances %>% 
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>% 
  mutate(mean_emp_euclidean = mean(emp_euclidean)) %>% 
  distinct(item_pairs, .keep_all = TRUE) %>% 
  ungroup() %>% 
  select(-c(item_pairs, subject_id, emp_euclidean))  %>% 
  mutate(itemA = factor(itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order))
  
# matrixify and plot
df_mean_distances_matrix <- df_mean_distances %>% 
  pivot_wider(names_from = itemB, values_from = mean_emp_euclidean) %>% 
  column_to_rownames("itemA")  
  
df_mean_distances_matrix <- df_mean_distances_matrix[desired_order, desired_order] #reorder matrix

diag(df_mean_distances_matrix) <- 0 #set diagonals 

# pdf(file = here(rdm_outputs_path, "figures/freesort/averageEuclideanRDM.pdf"), width = 8, height = 6)

corrplot(as.matrix(df_mean_distances_matrix), method = "color", is.corr = FALSE, 
         tl.col = "black", title = "Average Euclidean RDM", mar = c(0, 0, 2, 0),
         addgrid.col = "darkgray")

```

# Causal

## reshape data

```{r}
df_causation <- d %>%
    filter(trial_type == "causation") %>% 
    filter(!attention_check) %>% 
    select(all_of(causation_columns)) %>% 
    mutate(causal_judgments = 1 - response / 100) %>% 
    dplyr::select(-response) 

```

## plot subject-specific causal RDMs

```{r}
plot_causation_rdm <- function(data, subject, desired_order) {
  # Filter and process data for the specific subject
  df_matrix <- data %>%
    filter(subject_id == subject) %>%
    select(-subject_id) %>%
    select(itemA, itemB, causal_judgments) %>% 
    pivot_wider(names_from = itemB, values_from = causal_judgments) %>%
    column_to_rownames("itemA")
  
  # Reorder matrix and set diagonal to zero
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0
  
  # Plot the RDM
  corrplot(as.matrix(df_matrix), method = "color", is.corr = FALSE,
           tl.col = "black", title = paste("Causal RDM - ", subject),
           mar = c(0, 0, 2, 0), addgrid.col = "darkgray")
}

subject_ids <- unique(df_causation$subject_id)

#generate plots
# plots <- map(subject_ids, ~plot_causation_rdm(df_causation, .x, desired_order)) #run this to generate plots

# save plots with subject name as file name
 
# Generate unique subject IDs and save plots
new_names <- paste0("subj_", str_pad(seq_along(subject_ids), width = 2, pad = "0"))  # Generate new subject names

for (i in seq_along(subject_ids)) {  # Loop to save each plot as a separate PDF
  # Specify the path to save the PDF in the "figures" subfolder
  pdf(file = paste0(here(rdm_outputs_path), "/figures/causation/", new_names[i], ".pdf"), width = 8, height = 6)  # Open a PDF device
  plot_causation_rdm(df_causation, subject_ids[i], desired_order)  # Generate the plot
  dev.off()  # Close the device to save the file
}

```

## plot average causal RDM

```{r}
#normalize causality judgments
df_normalized <- df_causation %>% 
  mutate(item_pairs = str_c(itemA, itemB, sep = "-"))  %>%
  group_by(item_pairs) %>%
  mutate(mean_dissimilarity = mean(causal_judgments, na.rm = TRUE)) %>%  # Average across participants
  ungroup() 

#Min-Max Scaling 
df_normalized <- df_normalized %>%
  mutate(
    normalized_dissimilarity = (mean_dissimilarity - min(mean_dissimilarity)) /
                                (max(mean_dissimilarity) - min(mean_dissimilarity))
  )

df_rdm <- df_normalized %>%
  distinct(item_pairs, .keep_all = TRUE) %>% 
  select(itemA, itemB, normalized_dissimilarity) %>% 
  pivot_wider(names_from = itemB, values_from = normalized_dissimilarity) %>%
  column_to_rownames("itemA")  # Convert `itemA` to row names

df_rdm <- df_rdm[desired_order, desired_order]

# pdf(file = here(rdm_outputs_path, "figures/causation/averageCausalRDM.pdf"), width = 8, height = 6)  

corrplot(as.matrix(df_rdm), method = "color", is.corr = FALSE, tl.col = "black",
         title = "Average Causal RDM", mar = c(0, 0, 2, 0), addgrid.col = "darkgray")

```

# Correlations: Empirical euclidean RDMs

## read and compute correlations

```{r, message = FALSE}
d_all_theories <- read_csv(here("code/analysis/pilot/theoretical-rdm-data/d_all_theories.csv"))


# Function to extract upper triangular part as a vector
extract_upper_tri <- function(matrix) {
  matrix[upper.tri(matrix)]
}

d_correlations <- df_distances %>%
  left_join(d_all_theories, by = c("itemA", "itemB")) %>%
  left_join(df_causation, by = c("subject_id", "itemA", "itemB")) %>% 
  rename(`emp5_causal_judgments` = "causal_judgments") %>% 
  group_by(subject_id) %>%
  mutate(
      `1_Mind-Body-Action` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor1_Mind-Body-Action`),   #theoretical RDM
                          names_from = itemB, values_from = `theor1_Mind-Body-Action`) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
        
      `2_PerCog-ObjDir-Stimulus` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                            names_from = itemB, values_from = emp_euclidean) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`),     #theoretical RDM
                            names_from = itemB, values_from = `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order]     
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
    
    `3_Physical-Ethereal` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor3_Physical-Ethereal`),    #theoretical RDM
                          names_from = itemB, values_from = `theor3_Physical-Ethereal`) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
    
    `4_cosine_similarity` = {
              rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor4_cosine_similarity`),     #theoretical RDM
                          names_from = itemB, values_from = `theor4_cosine_similarity`) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
    
    `5_causal_judgments` = {
              rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `emp5_causal_judgments`),     #theoretical RDM
                          names_from = itemB, values_from = `emp5_causal_judgments`) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        }
    ) %>% 
  ungroup()


calculate_noise_ceiling <- function(data, subjects, desired_order, metric = "kendall") {
  # Initialize storage for correlations
  upper_bound <- numeric(length(subjects))
  lower_bound <- numeric(length(subjects))
  
  # Loop through each subject
  for (i in seq_along(subjects)) {
    subject <- subjects[i]
    
    # Subject's RDM
    subject_rdm <- data %>%
        filter(subject_id == subject) %>%
        select(subject_id, itemA, itemB, emp_euclidean) %>% 
        pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    subject_rdm <- subject_rdm[desired_order, desired_order]
    
    # Grand average RDM
    grand_avg_rdm <- data %>%
        select(subject_id, itemA, itemB, emp_euclidean) %>% 
        unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
        group_by(item_pairs) %>%
        mutate(emp_euclidean = mean(emp_euclidean)) %>%
        ungroup() %>% 
        distinct(item_pairs, .keep_all = TRUE) %>% 
        select(-c(item_pairs, subject_id)) %>% 
        pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    grand_avg_rdm <- grand_avg_rdm[desired_order, desired_order]
    
    # Other subjects' average RDM
    other_avg_rdm <- data %>%
      select(subject_id, itemA, itemB, emp_euclidean) %>% 
      filter(subject_id != "subj_01") %>%
      unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
      group_by(item_pairs) %>%
      mutate(emp_euclidean = mean(emp_euclidean)) %>%
      ungroup() %>% 
      distinct(item_pairs, .keep_all = TRUE) %>% 
      select(-c(item_pairs, subject_id)) %>% 
      pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
      column_to_rownames("itemA") %>%
      as.matrix()
    other_avg_rdm <- other_avg_rdm[desired_order, desired_order]
    
    # Convert to numeric vectors
    subject_upper <- as.numeric(subject_rdm[upper.tri(subject_rdm)])
    other_upper <- as.numeric(other_avg_rdm[upper.tri(other_avg_rdm)])
    grand_upper <- as.numeric(grand_avg_rdm[upper.tri(grand_avg_rdm)])
    
    # Compute correlations
    upper_bound[i] <- cor(subject_upper, other_upper, method = metric)
    lower_bound[i] <- cor(subject_upper, grand_upper, method = metric)
  }
  
  # Return average bounds
  list(
    upper_bound = mean(upper_bound, na.rm = TRUE),
    lower_bound = mean(lower_bound, na.rm = TRUE)
  )
}

# Compute noise ceiling for all subjects
subjects <- unique(d_correlations$subject_id)
noise_ceiling <- calculate_noise_ceiling(d_correlations, subjects, desired_order)

```

## plot correlations with noise ceiling

```{r, fig.width = 9}
# plot with noise ceiling
d_correlations %>%
  pivot_longer(cols = starts_with(c("1_","2_","3_","4_", "5_")),
               names_to = "correlation_type",
               values_to = "correlation_value") %>%
  ggplot(aes(x = reorder(correlation_type, correlation_value, FUN = function(x) -mean(x)), y = correlation_value)) +
  stat_summary(fun = "mean", geom = "bar", fill = "white", color = "black") +
  geom_point(aes(color = subject_id)) +
  geom_hline(yintercept = noise_ceiling$upper_bound, color = "blue", linetype = "dashed", linewidth = .5) +
  geom_hline(yintercept = noise_ceiling$lower_bound, color = "red", linetype = "dashed", linewidth = .5) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) +
  labs(x = "Theoretical RDMs", y = "Correlation with euclidean RDMs",
       title = "Correlation between theoretical and empirical RDMs",
       subtitle = paste0("Noise Ceiling: Upper: ", round(noise_ceiling$upper_bound, 3),
                         " | Lower: ", round(noise_ceiling$lower_bound, 3)))

```


# Correlations: Causal RDMS

## generate causal theories

```{r, echo = TRUE}
d_theo_causal_raw <- read_csv(here("code/analysis/pilot/theoretical-rdm-data/d_theories_categories.csv"))

d_theo_causal <- d_theo_causal_raw %>% 
  select(itemA, itemB, MindBodyAction_itemA, MindBodyAction_itemB) %>% 
  mutate(within_domain_int = case_when(
                          #intuitive bio
                          MindBodyAction_itemA == "bio" & MindBodyAction_itemB == "act" ~ 0, 
                          MindBodyAction_itemA == "act" & MindBodyAction_itemB == "bio" ~ 0,
                          MindBodyAction_itemA == "bio" & MindBodyAction_itemB == "bio" ~ 0,
                          #intuitive psych
                          MindBodyAction_itemA == "mind" & MindBodyAction_itemB == "act" ~ 0, 
                          MindBodyAction_itemA == "act" & MindBodyAction_itemB == "mind" ~ 0,
                          MindBodyAction_itemA == "mind" & MindBodyAction_itemB == "mind" ~ 0,
                          TRUE ~ 1),
         cross_domain_int = case_when(
                          MindBodyAction_itemA == "bio" & MindBodyAction_itemB == "act" ~ 0, 
                          MindBodyAction_itemA == "act" & MindBodyAction_itemB == "bio" ~ 0,
                          MindBodyAction_itemA == "bio" & MindBodyAction_itemB == "bio" ~ 0,
                          
                          MindBodyAction_itemA == "mind" & MindBodyAction_itemB == "act" ~ 0, 
                          MindBodyAction_itemA == "act" & MindBodyAction_itemB == "mind" ~ 0,
                          MindBodyAction_itemA == "mind" & MindBodyAction_itemB == "mind" ~ 0,
                          #cross-domain causation
                          MindBodyAction_itemA == "mind" & MindBodyAction_itemB == "bio" ~ 0, 
                          MindBodyAction_itemA == "bio" & MindBodyAction_itemB == "mind" ~ 0,
                          TRUE ~ 1)) %>% 
  select(-c("MindBodyAction_itemA", "MindBodyAction_itemB"))

  
```

## compute correlations

```{r}
d_causal_correlations <- df_distances %>%
  left_join(d_theo_causal, by = c("itemA", "itemB")) %>%
  left_join(df_causation, by = c("subject_id", "itemA", "itemB")) %>% 
  # rename(causal_judgments` = "causal_judgments") %>% 
  group_by(subject_id) %>%
  mutate(
      `1_emp_euclideans_upper` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, emp_euclidean),   # Construct theoretical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },

      `1_emp_euclideans_lower` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, emp_euclidean),   # Construct theoretical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        cor(rdm1[lower.tri(rdm1)], rdm2[lower.tri(rdm2)], method = "kendall")

        },

      `1_emp_euclideans_both` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, emp_euclidean),   # Construct theoretical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        # Flatten matrices and exclude NA values
        vec1 <- as.vector(rdm1)
        vec2 <- as.vector(rdm2)
        valid_indices <- !is.na(vec1) & !is.na(vec2)  # Exclude NA values
        
        cor(vec1[valid_indices], vec2[valid_indices], method = "kendall")
        },
      
      `2_within_domain_upper` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, within_domain_int),   # Construct theoretical RDM
                          names_from = itemB, values_from = within_domain_int) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")

        },
      
      `2_within_domain_lower` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, within_domain_int),   # Construct theoretical RDM
                          names_from = itemB, values_from = within_domain_int) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        cor(rdm1[lower.tri(rdm1)], rdm2[lower.tri(rdm2)], method = "kendall")

        },
      
      `2_within_domain_both` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, within_domain_int),   # Construct theoretical RDM
                          names_from = itemB, values_from = within_domain_int) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        # Flatten matrices and exclude NA values
        vec1 <- as.vector(rdm1)
        vec2 <- as.vector(rdm2)
        valid_indices <- !is.na(vec1) & !is.na(vec2)  # Exclude NA values
        
        cor(vec1[valid_indices], vec2[valid_indices], method = "kendall")

        },
      
      `3_cross_domain_upper` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, cross_domain_int),   # Construct theoretical RDM
                          names_from = itemB, values_from = cross_domain_int) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")

        },
      
      `3_cross_domain_lower` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, cross_domain_int),   # Construct theoretical RDM
                          names_from = itemB, values_from = cross_domain_int) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
        cor(rdm1[lower.tri(rdm1)], rdm2[lower.tri(rdm2)], method = "kendall")

        },
      
      `3_cross_domain_both` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, causal_judgments),     #empirical RDM
                          names_from = itemB, values_from = causal_judgments) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, cross_domain_int),   # Construct theoretical RDM
                          names_from = itemB, values_from = cross_domain_int) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        
                # Flatten matrices and exclude NA values
        vec1 <- as.vector(rdm1)
        vec2 <- as.vector(rdm2)
        valid_indices <- !is.na(vec1) & !is.na(vec2)  # Exclude NA values
        
        cor(vec1[valid_indices], vec2[valid_indices], method = "kendall")

        }) %>% 
  ungroup()


```

## plot correlations with empirical causal rdms

key

1. _upper: the upper tri of the causal rdm (since causal isn't symmetric, doing correlations for each causal tri separately. always using upper tri for empirical euclidean rdms since those are symmetric) 
2. _lower: lower tri  
3. _both: remove NAs, flatten into a vector, then compute correlations between the entire theoretical rdm and the causal rdms (NB: both is not avg of upper and lower)  
4. within_domain_: a classical causal rdm involving intuitive theories within domains. mind causes act (and vice versa), bio causes act (and vice versa), mind causes mind, and bio causes bio.  
5. cross_domain_: classical causal rdm per point 4 PLUS causal relation between bio and mind (bio cause mind and vice versa) 

```{r, fig.width = 12}
# plot with noise ceiling
d_causal_correlations %>%
  pivot_longer(cols = starts_with(c("1_", "2_", "3_")),
               names_to = "correlation_type",
               values_to = "correlation_value") %>%
  mutate(category = case_when(str_starts(correlation_type, "1_") ~ "empirical euclidean",
                              str_starts(correlation_type, "2_") ~ "with_dom int theory",
                              str_starts(correlation_type, "3_") ~ "cross_dom int theory")) %>% 
  ggplot(aes(x = reorder(correlation_type, correlation_value, FUN = function(x) -mean(x)), y = correlation_value, fill = category)) +
  stat_summary(fun = "mean", geom = "bar", color = "black") +
  # stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.5) +
  geom_point(aes(color = subject_id)) +
  geom_hline(yintercept = noise_ceiling$upper_bound, color = "blue", linetype = "dashed", linewidth = .5) +
  geom_hline(yintercept = noise_ceiling$lower_bound, color = "red", linetype = "dashed", linewidth = .5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) +
  labs(x = "Theoretical RDMs", y = "Correlation with causal RDMs",
       title = "Correlation between theoretical and empirical RDMs",
       subtitle = paste0("Noise Ceiling: Upper: ", round(noise_ceiling$upper_bound, 3),
                         " | Lower: ", round(noise_ceiling$lower_bound, 3)))


```


still to do: causal rdms where everything can cause everything except for certain conditions..  

- There are limits to what can cause perception (seeing or hearing something) - at least directly. For instance, mental and bodily events can’t literally cause you to see something. But moving one’s body around can.    
- Certain physiological events limit what the body can do - e.g. feeling sick and hungry tends to not lead to high-intensity actions like jumping, kicking, walking. Energy flows downhill.  
- Certain actions constrain what next actions are possible, e.g. resting actions like sitting down tend not to cause high energy actions.  
- ??? Perceptual events (seeing or hearing) cannot directly cause physiological states like being tired, being in pain - but cognitive events can.   

