---
title: "causation"
output: html_document
date: "2024-10-28"
---

# ReadMe

This is the analysis for study 2 of the MBC project, alias the intervention study.  
The study had N = 100, two between subjects conditions: Intervention and Similarity, and 15 trials per condition.
The question: are subjects in the Intervention condition more likely to select the causal item, and subjects in the Similarity condition likely to select the similar item.

# Set options

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(tidyverse, lme4, here)
sessionInfo()

```

# Read data

```{r, warnings = FALSE}
base_path <- here("code", "analysis", "pilot")
data_path <- paste0(base_path, "/data-intervention/pilot06_11425_n100-intervention-100")
snapshots_input_path <- paste0(base_path, "/snapshots_study1_final_analysis")

desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

file_paths <- list.files(path = data_path, pattern = "^2025-", full.names = TRUE)
d <- map_df(file_paths, ~ read_csv(.x) %>% mutate(filename = basename(.x)))

```

# Wrangle data

```{r}
 subjects_to_exclude <- d %>%
  filter(condition == "attention_check") %>% 
  rowwise() %>% 
  mutate(correct_attn_check = case_when(
    grepl(correct_response_attn, selected_choice) ~ "yes",
    TRUE ~ "no"
  )) %>% 
  filter(correct_attn_check == "no") %>% 
  pull(filename)

dat <- d %>% 
  filter(!filename %in% subjects_to_exclude) %>% #exclude 1 subject who failed attn check
  filter(condition %in% c("intervention", "similarity")) %>%
  group_by(filename) %>%
  mutate(subject_id = paste0("subj_", cur_group_id())) %>%
  ungroup() %>% 
  rowwise() %>%
  mutate(response = case_when(
    grepl(causal_itemA, selected_choice) ~ "causal",
    grepl(similarity_itemA, selected_choice) ~ "similarity",
    TRUE ~ "other"
  )) %>% 
  ungroup() %>% 
  select(subject_id, condition, itemB, causal_itemA, similarity_itemA, response) %>% 
  mutate(causal_itemA = factor(causal_itemA, levels = desired_order),
         similarity_itemA = factor(similarity_itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(domain_itemB = case_when(itemB == "hear something" ~ "mind", #labels for first item of pair
                                 itemB == "choose what to do" ~ "mind",
                                 itemB == "remember something" ~ "mind",
                                 itemB == "think about something" ~ "mind",
                                 itemB == "see something" ~ "mind",
                                 itemB == "reach for something" ~ "action",
                                 itemB == "sit down" ~ "action", 
                                 itemB == "jump up and down" ~ "action",
                                 itemB == "kick something" ~ "action",
                                 itemB == "take a walk" ~ "action",
                                 itemB == "get tired" ~ "bio", 
                                 itemB == "become hungry" ~ "bio",
                                 itemB == "feel scared"~ "bio",
                                 itemB == "experience pain"~ "bio",
                                 itemB == "get sick"~ "bio"),
         domain_causalitemA = case_when(causal_itemA == "hear something" ~ "mind", 
                                 causal_itemA == "choose what to do" ~ "mind",
                                 causal_itemA == "remember something" ~ "mind",
                                 causal_itemA == "think about something" ~ "mind",
                                 causal_itemA == "see something" ~ "mind",
                                 causal_itemA == "reach for something" ~ "action",
                                 causal_itemA == "sit down" ~ "action", 
                                 causal_itemA == "jump up and down" ~ "action",
                                 causal_itemA == "kick something" ~ "action",
                                 causal_itemA == "take a walk" ~ "action",
                                 causal_itemA == "get tired" ~ "bio", 
                                 causal_itemA == "become hungry" ~ "bio",
                                 causal_itemA == "feel scared"~ "bio",
                                 causal_itemA == "experience pain"~ "bio",
                                 causal_itemA == "get sick"~ "bio"),
          domain_similarityitemA = case_when(similarity_itemA == "hear something" ~ "mind", 
                                 similarity_itemA == "choose what to do" ~ "mind",
                                 similarity_itemA == "remember something" ~ "mind",
                                 similarity_itemA == "think about something" ~ "mind",
                                 similarity_itemA == "see something" ~ "mind",
                                 similarity_itemA == "reach for something" ~ "action",
                                 similarity_itemA == "sit down" ~ "action", 
                                 similarity_itemA == "jump up and down" ~ "action",
                                 similarity_itemA == "kick something" ~ "action",
                                 similarity_itemA == "take a walk" ~ "action",
                                 similarity_itemA == "get tired" ~ "bio", 
                                 similarity_itemA == "become hungry" ~ "bio",
                                 similarity_itemA == "feel scared"~ "bio",
                                 similarity_itemA == "experience pain"~ "bio",
                                 similarity_itemA == "get sick"~ "bio"))


```

# Plot

## overall barplot

```{r}
dat_summarized <- dat %>% 
  group_by(condition) %>%
  mutate(total = n()) %>% 
  group_by(condition, response) %>%
  summarize(
    count = n(),  
    total = first(total), 
    mean_proportion = count / total, 
    se = sqrt(mean_proportion * (1 - mean_proportion) / total),  
    .groups = "drop")

#stacked
ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw()

#unstacked
ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(
      ymin = mean_proportion - se,
      ymax = mean_proportion + se
    ),
    position = position_dodge(0.9),  #aign error bars with dodged bars
    width = 0.2
  ) +
  theme_bw()

```

## item level, stacked

```{r, fig.width = 12}
dat_summarized_by_item <- dat %>% 
  group_by(condition, itemB) %>%
  mutate(total = n()) %>%  # Total responses per condition
  group_by(condition, itemB, response) %>%
  mutate(
    count = n(),  # Count responses for each response type
    total = first(total),  # Total responses for the condition
    mean_proportion = count / total,  # Proportion of response within condition
    se = sqrt(mean_proportion * (1 - mean_proportion) / total)) %>% 
  distinct(condition, itemB, .keep_all = TRUE) %>% 
  ungroup()

#set x axis colors
itemB_colors <- dat_summarized_by_item %>%
  distinct(itemB, domain_itemB) %>%
  mutate(color = case_when(
    domain_itemB == "bio" ~ "green",
    domain_itemB == "mind" ~ "blue",
    domain_itemB == "action" ~ "red"
  )) %>%
  pull(color, name = itemB)

p <- dat_summarized_by_item %>% 
  ggplot(aes(x =itemB, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(dat_summarized_by_item$itemB)])) +
  facet_wrap(~condition) 

p

ggsave(
  filename = "figures/study2/p_study2_item-level_stacked.pdf",
  plot = p,          
  device = "pdf",           
  #path = "path/to/directory", #Optional: specify a directory
  width = 12,               
  height = 6,             
  units = "in"              
)

```

## item-level, unstacked with error bars

```{r, fig.width = 12}
dat_summarized_by_item %>% 
  ggplot(aes(x =itemB, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(
    aes(
      ymin = mean_proportion - se,
      ymax = mean_proportion + se
    ),
    position = position_dodge(0.9),  #align error bars with dodged bars
    width = 0.2
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(dat_summarized_by_item$itemB)])) +
  facet_wrap(~condition) 

```

# Modelling

## effect of condition

Todo: rename mean_emp_euclidean and mean_causal_dissimilarity to causal and freesort distance in all scripts!

```{r}
d_model <- dat %>% 
  mutate(response = as.factor(response)) %>% 
  mutate(subject_id = as.factor(subject_id)) %>% 
  mutate(responses = ifelse(response == "causal", 1, 0)) %>% 
  mutate(condition = as.factor(condition))
  
#model
overall_model <- glmer(responses ~ condition + (1|subject_id), data = d_model, family = binomial(link = 'logit'))
summary(overall_model)

```

## effect of causal and freesort distance in causal condition

```{r}
d_freesort_and_causal_distances <- read.csv(paste0(snapshots_input_path, "/diffs.csv")) %>% 
  mutate(itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(itemA = factor(itemA, levels = desired_order)) 

causal_condition <- d_model %>% 
  select(-similarity_itemA) %>% 
    filter(condition == "intervention") %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean) %>% 
                rename("causal_itemA" = itemA,
                       "causal_distance" = mean_causal_dissimilarity,
                       "freesort_distance" = mean_emp_euclidean),
              by = c("itemB", "causal_itemA")) 

similarity_condition <- d_model %>% 
  select(-causal_itemA) %>% 
  filter(condition == "similarity") %>% 
  left_join(d_ratings %>% 
                select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean) %>% 
                rename("similarity_itemA" = itemA,
                       "causal_distance" = mean_causal_dissimilarity,
                       "freesort_distance" = mean_emp_euclidean),
              by = c("itemB", "similarity_itemA"))

#Modelling both predictors 
both_predictors_model <- glmer(responses ~ causal_distance + freesort_distance + (1|subject_id), data = causal_condition, family = binomial(link = "logit"))
summary(both_predictors_model)

fixed_effects <- fixef(both_predictors_model) #log odds
odds_ratios <- exp(fixed_effects)  #convert log-odds to odds ratios
probabilities <- 1 / (1 + exp(-fixed_effects)) #convert log odds to probability

cat("\nBoth predictors model [intercept] [causal_distance] [freesort_distance]:\n ")
cat("Fixed effects (log odds): ", fixed_effects, "\n")
cat("Fixed effects (odds ratios): ", odds_ratios, "\n")
cat("Fixed effects (probabilities, causal_distance = 0):", probabilities, "\n")
#nb: fixed effects are intercept and slope. slope is change log odds of outcome for a unit increase in predictor

```

## effect of causal and freesort distance in similarity condition

```{r}
# Modelling predictors jointly
both_predictors_model <- glmer(responses ~ causal_distance + freesort_distance + (1|subject_id), data = similarity_condition, family = binomial(link = "logit"))
summary(both_predictors_model)

fixed_effects <- fixef(both_predictors_model) #log odds
odds_ratios <- exp(fixed_effects)  #convert log-odds to odds ratios
probabilities <- 1 / (1 + exp(-fixed_effects)) #convert log odds to probability

cat("\nBoth predictors model [intercept] [causal_distance] [freesort_distance]:\n ")
cat("Fixed effects (log odds): ", fixed_effects, "\n")
cat("Fixed effects (odds ratios): ", odds_ratios, "\n")
cat("Fixed effects (probabilities, causal_distance = 0):", probabilities, "\n")

```

## predicting effect sizes (i.e. proportion differences) based on differences in rating scores

### first get predicted score differences

```{r}
#getting predictions for causal condition (most causal item)
diffs_causal <-d_freesort_and_causal_distances %>%
  group_by(itemB, domain_itemA) %>%
  slice(which.max(euclidean_minus_causal)) %>% 
  ungroup() %>% 
  group_by(itemB) %>% 
  slice(which.max(euclidean_minus_causal)) %>% 
  ungroup() %>% 
  rename("causal_itemA" = itemA,
         "domain_causal_itemA" = domain_itemA,
         "causal_item_euc_minus_causal" = euclidean_minus_causal) %>%
  select(causal_itemA, itemB, domain_causal_itemA, domain_itemB, causal_item_euc_minus_causal)

#get predictions for causal condition (most causal item)
diffs_similarity <- d_freesort_and_causal_distances %>%
  group_by(itemB, domain_itemA) %>%
  slice(which.min(euclidean_minus_causal)) %>% 
  ungroup() %>% 
  mutate(same_domain = ifelse(domain_itemA == domain_itemB, TRUE, FALSE)) %>% 
  filter(same_domain == TRUE) %>% 
  ungroup() %>% 
  rename("similarity_itemA" = itemA,
       "domain_similarity_itemA" = domain_itemA,
       "similarity_item_euc_minus_causal" = euclidean_minus_causal) %>% 
  select(similarity_itemA, itemB, domain_similarity_itemA, domain_itemB, similarity_item_euc_minus_causal)

d_diffs <- diffs_causal %>% 
  left_join(diffs_similarity %>% 
              select(everything()), 
              by = c("itemB", "domain_itemB")) %>% 
  mutate(diff = causal_item_euc_minus_causal - similarity_item_euc_minus_causal)

```

### then plot predicted score differences

```{r}
dat_combined <- dat_summarized_by_item %>%
  group_by(condition, itemB) %>%
  reframe(
    prop_causal = mean(mean_proportion[response == "causal"], na.rm = TRUE),
    prop_similar = mean(mean_proportion[response == "similarity"], na.rm = TRUE),
    prop_causal = replace_na(prop_causal, 0),  #fill missing values with 0: (for some items all subj choose 0, causing NaNs or their proportions)
    prop_similar = replace_na(prop_similar, 0), 
    observed_diff = case_when(
      condition == "intervention" ~ prop_causal - prop_similar,  
      condition == "similarity" ~ prop_similar - prop_causal  
    )
  ) %>%
  left_join(d_diffs %>% select(itemB, diff), by = "itemB")  # merge in predicted differences

```

## correlation bar plot

something's wrong with remember and think something -- look into it!!
```{r}
itemB_colors <- dat_summarized_by_item %>%
  distinct(itemB, domain_itemB) %>%
  mutate(color = case_when(
    domain_itemB == "bio" ~ "green",
    domain_itemB == "mind" ~ "blue",
    domain_itemB == "action" ~ "red"
  )) %>%
  pull(color, name = itemB)

barplot <- dat_combined %>%
  ggplot(aes(x = itemB)) +
  geom_bar(aes(y = observed_diff), stat = "identity", fill = "blue", alpha = 0.6) +
  geom_point(aes(y = diff), color = "red", size = 3) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(diffs_causal$itemB)])
  ) +
  labs(
    title = "Observed vs Predicted Proportion Differences by Item",
    x = "Item B",
    y = "observed proportion differences"
  ) +
  facet_wrap(~condition)

barplot

```


## correlation scatter plot 

may be worse because of infinity values for remember and think about something....look into it!

```{r}
# Scatter plot with correlation for each condition
scatter_plot <- dat_combined %>%
  ggplot(aes(x = diff, y = observed_diff, color = condition)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  theme_bw() +
  facet_wrap(~condition) +
  labs(
    title = "Observed vs Predicted Differences by Condition",
    x = "Predicted proportion difference",
    y = "Observed proportion difference",
    color = "Condition"
  )

scatter_plot
```

### quantitative correlations 

```{r}
overall_correlations <- dat_combined %>%
  summarize(
    correlation = cor(diff, observed_diff, use = "complete.obs"),
    p_value = cor.test(diff, observed_diff, use = "complete.obs")$p.value,
    .groups = "drop"
  )

print(overall_correlations)

condition_level_correlations <- dat_combined %>%
  group_by(condition) %>%
  summarize(
    correlation = cor(diff, observed_diff, use = "complete.obs"),
    p_value = cor.test(diff, observed_diff, use = "complete.obs")$p.value,
    .groups = "drop"
  )

print(condition_level_correlations)

```


# Plotting logistic regression models

## Intervention condition


```{r}
#create a new dataset for prediction
prediction_data <- data.frame(
  causal_distance = seq(min(causal_condition$causal_distance), 
                        max(causal_condition$causal_distance), 
                        length.out = 100)
)

#add predicted probabilities to the dataset
prediction_data$predicted_prob <- predict(
  glm(formula = responses ~ causal_distance, family = binomial(link = "logit"), data = causal_condition),
  newdata = prediction_data,
  type = "response"
)

#plot 1: causal distance
ggplot(data = causal_condition, aes(x = causal_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving a Causal Response in Intervention Condition",
    y = "Probability",
    color = "Actual Response"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

#plot 2: freesort distance
ggplot(data = causal_condition, aes(x = freesort_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving a Causal Response in Intervention Condition",
    y = "Probability",
    color = "Actual Response"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

```

## similarity condition

```{r}
#plot 1: freesort distance
ggplot(data = similarity_condition, aes(x = freesort_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving Causal Response in Similarity Condition",
    y = "Prob",
    color = "observed response"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

#plot 2: causal distance
ggplot(data = similarity_condition, aes(x = causal_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving Causal Response in Similarity Condition",
    y = "Prob",
    color = "observed responses"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

```













