---
title: "mbc-freesort"
output: html_document
date: "2024-11-12"
---

```{r setup, include=FALSE}
library(pacman)
pacman::p_load(tidyverse,
               ggforce,
               conflicted,
               here,
               cowplot,
               lme4,
               lmerTest,
               effects,
               performance,
               purrr,
               readr,
               igraph,
               ggraph,
               tidygraph,
               visNetwork,
               jsonlite,
               corrplot,
               stringr,
               scales)

library(plotly) #for some reason this never loads via pacman
walk(c("select", "filter", "rename", "mutate", "summarize"), ~ conflict_prefer(.x, "dplyr"))
conflicts_prefer(plotly::layout)
conflicts_prefer(here::here)
theme_set(theme_bw())
sessionInfo()

Sys.setenv(RETICULATE_PYTHON = "/Users/jouta1/torch-env/bin/python") #because default r-reticulate is linked to python 3.2, which does not support torch, so we create a new environment which we load with python 3.1 and torch
library(reticulate)
options(reticulate.verbose = TRUE)
py_config()
py_run_string("import tensorflow as tf; print(tf.__version__)")
library(text)
library(stringr)
library(patchwork)#for combining plots
library(forcats) #for reordering factors

```

# Read 

```{r}
directory_path <- here("code", "analysis", "pilot", "data-freesort")
file_paths <- list.files(path = directory_path, full.names = TRUE)
d <- map_df(file_paths, ~ read_csv(.x) %>% mutate(filename = basename(.x)))

```

# Extract relevant columns

```{r}
items <- c("choose what to do", "reach for something", "feel scared",  #items in exact order of img1.png to img15.png
           "experience pain", "get sick", "get tired", "become hungry", 
           "see something", "think about something", "hear something", 
           "remember something", "jump up and down", "kick something", 
           "take a walk", "sit down")

desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

df <- d %>% 
  relocate(mturk_participant_id, .before = "success") %>%
  filter(trial_type == "free-sort") %>%
  select(mturk_participant_id, final_locations) %>%
  mutate(parsed_location = map(final_locations, fromJSON)) %>% #extract data saved as json
  unnest(parsed_location) %>% # Expand parsed JSON 
  mutate(img_number = as.numeric(str_extract(src, "\\d+")), #attach items
         item = items[img_number]) %>%
  select(-c("final_locations", "src", "img_number")) %>% 
  rename("subject_id" = mturk_participant_id) %>% 
  mutate(subject_id = case_when(subject_id == "1A34FA8D1D1947E7B5F62D1717AE2C24" ~ "subj_01",
                                subject_id == "36750B8567FA4C1AA9B9C55C1DB3B7EE" ~ "subj_02",
                                subject_id == "637A401ADBB14DA1887A79D522A7071A" ~ "subj_03",
                                subject_id == "79D36E8ABBDA48B2912AE910028C2A3F" ~ "subj_04",
                                subject_id == "99F3A46BBF8D4B72A98A8CE941CBF67A" ~ "subj_05",
                                subject_id == "A0521F3CB2AF452C9817A55859770623" ~ "subj_06",
                                subject_id == "BD1C3DB4C40D4234AF38A82DE44A1B06" ~ "subj_07",
                                subject_id == "DAEFF5D302D94311B5DE8A4787CE2CA8" ~ "subj_08",
                                subject_id == "DDAB249809254CBFB91F03D45E6A2F9F" ~ "subj_09",
                                subject_id == "FE9EB599CD2142EA912913DE16ED9161" ~ "subj_10"))

```

# Visualize raw responses

## plot individual subjects

```{r, fig.width = 12}
# pdf("my_plot_raw.pdf", width = 12, height = 7)  

df %>% 
  mutate(y = 500 - y) %>% #invert y axis
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  geom_text(aes(label = item), size = 3, vjust = -.2) +
  facet_wrap(~subject_id)

```

## plot average raw responses

```{r}
# pdf("my_plot_average.pdf", width = 7, height = 4)  # Width and height are in inches

df %>% 
  mutate(y = 500 - y) %>% 
  group_by(item) %>% 
  mutate(mean_x = mean(x),
            mean_y = mean(y)) %>% 
  ungroup() %>% 
  distinct(item, .keep_all = TRUE) %>% 
  select(-subject_id) %>% 
  ggplot(aes(x = mean_x, y = mean_y)) +
  geom_point() +
  geom_text(aes(label = item), vjust = -1) +
  labs(title = "Average Raw Responses")

```


# Visualize empirical euclidean distances

## reshape data

```{r}
#get item pairs per person
df_distances <- df %>%
  select(subject_id, item) %>%
  group_by(subject_id) %>%
  summarize(pairs = list(expand.grid(itemA = item, itemB = item, stringsAsFactors = FALSE))) %>%
  unnest(pairs) %>%
  filter(itemA != itemB) %>% #remove self-pairs
  left_join(df, by = c("subject_id", "itemA" = "item")) %>% #append itemA positions
  rename(x_itemA = x, y_itemA = y) %>% 
  left_join(df, by = c("subject_id", "itemB" = "item")) %>% #append itemB positions
  rename(x_itemB = x, y_itemB = y) %>% 
  group_by(subject_id) %>% 
  mutate(euclidean_distance = sqrt((x_itemA - x_itemB)^2 + (y_itemA - y_itemB)^2),
         emp_euclidean = ((euclidean_distance - min(euclidean_distance)) /
                                   (max(euclidean_distance) - min(euclidean_distance)))) %>% #normalizing euclidean distances
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  ungroup() %>%
  mutate(itemA = factor(itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order))

```

## plot subject-specific rdms

```{r}
plot_rdm <- function(data, subject, desired_order) { # Function to create RDM plot for a single subject
  df_matrix <- data %>%
    filter(subject_id == subject) %>%
    select(-subject_id) %>%
    select(itemA, itemB, emp_euclidean) %>% 
    pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
    column_to_rownames("itemA")
  
  df_matrix <- df_matrix[desired_order, desired_order] #reorder matrix
  diag(df_matrix) <- 0  # Set diagonal to 0
  
  corrplot(as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = paste("RDM - ", subject),
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
}

subject_ids <- unique(df_distances$subject_id)

#generate plots
plots <- map(subject_ids, ~plot_rdm(df_distances, .x, desired_order)) #run this entire line of code

# save plots with subject name
subject_ids <- unique(df_distances$subject_id)
new_names <- paste0("subj_", str_pad(seq_along(subject_ids), width = 2, pad = "0")) # Generate new subject names
for (i in seq_along(subject_ids)) { # Loop to save each plot as a separate PDF
  pdf(paste0(new_names[i], ".pdf"), width = 8, height = 6)  # Open a PDF device
  plot_rdm(df_distances, subject_ids[i], desired_order)  # Generate the plot
  dev.off()  # Close the device to save the file
}

```

## plot averaged RDM over all subjects 

```{r}
df_mean_distances <- df_distances %>% 
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>% 
  mutate(mean_emp_euclidean = mean(emp_euclidean)) %>% 
  distinct(item_pairs, .keep_all = TRUE) %>% 
  ungroup() %>% 
  select(-c(item_pairs, subject_id, emp_euclidean))  %>% 
  mutate(itemA = factor(itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order))
  
# matrixify and plot
df_mean_distances_matrix <- df_mean_distances %>% 
  pivot_wider(names_from = itemB, values_from = mean_emp_euclidean) %>% 
  column_to_rownames("itemA")  
  
df_mean_distances_matrix <- df_mean_distances_matrix[desired_order, desired_order] #reorder matrix

diag(df_mean_distances_matrix) <- 0 #set diagonals 

corrplot(as.matrix(df_mean_distances_matrix), method = "color", is.corr = FALSE, 
         tl.col = "black", title = "AVERAGE RDM", mar = c(0, 0, 2, 0),
         addgrid.col = "darkgray")

```

# Visualize Theoretical RDMs

## First generate the theoretical data 

```{r}
d_theories <- expand.grid(itemA = desired_order, itemB = desired_order) %>% #expand grid
  filter(itemA != itemB) %>%  # Remove self-pairs
  
  #theory 1: (3 groups) mind, body, action
  mutate(MindBodyAction_itemA = case_when(itemA == "hear something" ~ "mind", #labels for first item of pair
                               itemA == "choose what to do" ~ "mind",
                               itemA == "remember something" ~ "mind",
                               itemA == "think about something" ~ "mind",
                               itemA == "see something" ~ "mind",
                               itemA == "reach for something" ~ "act",
                               itemA == "sit down" ~ "act", 
                               itemA == "jump up and down" ~ "act",
                               itemA == "kick something" ~ "act",
                               itemA == "take a walk" ~ "act",
                               itemA == "get tired" ~ "bio", 
                               itemA == "become hungry" ~ "bio",
                               itemA == "feel scared"~ "bio",
                               itemA == "experience pain"~ "bio",
                               itemA == "get sick"~ "bio"),
         MindBodyAction_itemB = case_when(itemB == "hear something" ~ "mind", #labels for second item of pair
                               itemB == "choose what to do" ~ "mind",
                               itemB == "remember something" ~ "mind",
                               itemB == "think about something" ~ "mind",
                               itemB == "see something" ~ "mind",
                               itemB == "reach for something" ~ "act",
                               itemB == "sit down" ~ "act", 
                               itemB == "jump up and down" ~ "act",
                               itemB == "kick something" ~ "act",
                               itemB == "take a walk" ~ "act",
                               itemB == "get tired" ~ "bio", 
                               itemB == "become hungry" ~ "bio",
                               itemB == "feel scared"~ "bio",
                               itemB == "experience pain"~ "bio",
                               itemB == "get sick"~ "bio")) %>% 
  
  #theory 2: (6 groups) perceptual, cognition, object directed, non-object directed, bodily tied to simulation, bodily not tied to stimulation
   mutate(PerCogOdActSpActBstimBnstim_itemA = case_when(itemA == "hear something" ~ "per",
                               itemA == "see something" ~ "per",
                               itemA == "choose what to do" ~ "cog",
                               itemA == "remember something" ~ "cog",
                               itemA == "think about something" ~ "cog",
                               itemA == "reach for something" ~ "odAct",
                               itemA == "sit down" ~ "SpAct", 
                               itemA == "jump up and down" ~ "SpAct",
                               itemA == "kick something" ~ "odAct",
                               itemA == "take a walk" ~ "SpAct",
                               itemA == "get tired" ~ "BnS", 
                               itemA == "become hungry" ~ "BnS",
                               itemA == "feel scared"~ "BS",
                               itemA == "experience pain"~ "BS",
                               itemA == "get sick"~ "BnS",),
         PerCogOdActSpActBstimBnstim_itemB = case_when(itemB == "hear something" ~ "per",
                               itemB == "see something" ~ "per",
                               itemB == "choose what to do" ~ "cog",
                               itemB == "remember something" ~ "cog",
                               itemB == "think about something" ~ "cog",
                               itemB == "reach for something" ~ "odAct",
                               itemB == "sit down" ~ "SpAct", 
                               itemB == "jump up and down" ~ "SpAct",
                               itemB == "kick something" ~ "odAct",
                               itemB == "take a walk" ~ "SpAct",
                               itemB == "get tired" ~ "BnS", 
                               itemB == "become hungry" ~ "BnS",
                               itemB == "feel scared"~ "BS",
                               itemB == "experience pain"~ "BS",
                               itemB == "get sick"~ "BnS")) %>% 

      #theory 3: (2 groups) physical and non-physical events:
      mutate(PhysicalEthereal_itemA = case_when(itemA == "hear something" ~ "ethereal",
                               itemA == "see something" ~ "ethereal",
                               itemA == "choose what to do" ~ "ethereal",
                               itemA == "remember something" ~ "ethereal",
                               itemA == "think about something" ~ "ethereal",
                               itemA == "reach for something" ~ "phys",
                               itemA == "sit down" ~ "phys", 
                               itemA == "jump up and down" ~ "phys",
                               itemA == "kick something" ~ "phys",
                               itemA == "take a walk" ~ "phys",
                               itemA == "get tired" ~ "phys", 
                               itemA == "become hungry" ~ "phys",
                               itemA == "feel scared"~ "phys",
                               itemA == "experience pain"~ "phys",
                               itemA == "get sick"~ "phys"),
             PhysicalEthereal_itemB = case_when(itemB == "hear something" ~ "ethereal",
                               itemB == "see something" ~ "ethereal",
                               itemB == "choose what to do" ~ "ethereal",
                               itemB == "remember something" ~ "ethereal",
                               itemB == "think about something" ~ "ethereal",
                               itemB == "reach for something" ~ "phys",
                               itemB == "sit down" ~ "phys", 
                               itemB == "jump up and down" ~ "phys",
                               itemB == "kick something" ~ "phys",
                               itemB == "take a walk" ~ "phys",
                               itemB == "get tired" ~ "phys", 
                                itemB == "become hungry" ~ "phys",
                               itemB == "feel scared"~ "phys",
                               itemB == "experience pain"~ "phys",
                               itemB == "get sick"~ "phys",)) %>% 
  
  #scoring the items
  mutate("theor1_Mind-Body-Action" = ifelse(MindBodyAction_itemA == MindBodyAction_itemB, 0, 1),
         "theor2_PerCog-OdActSpAct-BodyStimBodynoStim" = ifelse(PerCogOdActSpActBstimBnstim_itemA == PerCogOdActSpActBstimBnstim_itemB, 0, 1),
         "theor3_Physical-Ethereal" = ifelse(PhysicalEthereal_itemA == PhysicalEthereal_itemB, 0, 1)) %>% 
  select(-c("MindBodyAction_itemA", "MindBodyAction_itemB", 
            "PerCogOdActSpActBstimBnstim_itemA","PerCogOdActSpActBstimBnstim_itemB", 
            "PhysicalEthereal_itemA", "PhysicalEthereal_itemB"))

```


## then add USE embeddings and cosine similarities

### create and Save USE embeddings

```{r}
# Pyhton
# first open terminal and activate virtual environment:
#source /Users/jouta1/torch-env/bin/activate #enter this on terminal
#rm -rf /var/folders/pf/zr1pbgy56sgf04934jmg2z6msvz06s/T/tfhub_modules #run this to clear cached model files
#then run this code:

# Step 1: Extract unique items
# unique_items <- d_theories %>%
#   select(itemA) %>%
#   distinct() %>% 
#   pull()
# 
# # Step 2: Compute embeddings using USE
# hub <- import("tensorflow_hub")
# model <- hub$load("https://tfhub.dev/google/universal-sentence-encoder/4")
# 
# # Create a named list of embeddings
# item_embeddings <- lapply(unique_items, function(item) {
#   embedding <- model(list(item))$numpy()
#   as.numeric(embedding)
# })
# names(item_embeddings) <- unique_items
# 
# # Convert embeddings into a data frame for easier merging
# embedding_df <- data.frame(
#   item = names(item_embeddings),
#   embedding = I(item_embeddings)
# )
# 
# saveRDS(embedding_df, file = "embeddings/item_embeddings.rds") #save embeddings
```


### load USE embeddings

```{r}
embedding_df <- readRDS("embeddings/item_embeddings.rds") #read embeddings

# Step 3: Merge embeddings back into the long data frame
# Add embeddings for itemA and itemB
d_theories_with_embeddings <- d_theories %>%
  left_join(embedding_df, by = c("itemA" = "item")) %>%
  rename(embeddingA = embedding) %>%
  left_join(embedding_df, by = c("itemB" = "item")) %>%
  rename(embeddingB = embedding)

# Compute cosine similarity for each row
d_all_theories <- d_theories_with_embeddings %>%
  rowwise() %>% 
  mutate(cosine_similarity = sum(unlist(embeddingA) * unlist(embeddingB)) /
                 (sqrt(sum(unlist(embeddingA)^2)) * sqrt(sum(unlist(embeddingB)^2)))) %>%
  ungroup() %>% 
  rowwise() %>% 
  mutate(cosine_similarity = 1 - cosine_similarity) %>% #turning cosine similarity to a disimilarity measure. For some reason high cosine similarity means disimilar
  ungroup() %>% 
  select(-c("embeddingA", "embeddingB")) %>% 
  rename("theor4_cosine_similarity" = cosine_similarity) %>% 
  ungroup() 

```


## Plot theoretical proposals

```{r}
plot_rdm_theory <- function(data, column_name, desired_order, plot_title) {
  # Prepare data and reorder
  df_matrix <- data %>%
    mutate(
      itemA = factor(itemA, levels = desired_order),
      itemB = factor(itemB, levels = desired_order)
    ) %>%
    select(itemA, itemB, all_of(column_name)) %>%
    pivot_wider(names_from = itemB, values_from = all_of(column_name)) %>%
    column_to_rownames("itemA")
  
  # Reorder rows and columns
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0 # Set diagonal to 0
  
  # Plot RDM
  corrplot(
    as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = plot_title,
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
}

# Define desired orders
mba_order <- c("see something", "hear something", "choose what to do", 
               "remember something", "think about something", "reach for something", 
               "sit down", "jump up and down", "kick something", "take a walk", 
               "get tired", "become hungry", "feel scared", "experience pain", "get sick")

perccog_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "kick something", "sit down", "jump up and down",  "take a walk", 
                   "get tired", "become hungry", "get sick", "feel scared", "experience pain")

physethereal_order <- c("see something", "hear something", "choose what to do", 
                        "remember something", "think about something", "reach for something", 
                        "kick something", "sit down", "jump up and down",  "take a walk", 
                        "get tired", "become hungry", "get sick", "feel scared", "experience pain")

# Plot RDMs for each theory
plot_rdm_theory(d_all_theories, "theor4_cosine_similarity", mba_order, "1. Cosine Distance - RDM")
plot_rdm_theory(d_all_theories, "theor1_Mind-Body-Action", mba_order, "2. Mind, Body & Action - RDM")
plot_rdm_theory(d_all_theories, "theor2_PerCog-OdActSpAct-BodyStimBodynoStim", perccog_order, "3. Perception, Cognitive et al. - RDM")
plot_rdm_theory(d_all_theories, "theor3_Physical-Ethereal", physethereal_order, "4. Physical & Ethereal - RDM")

```

```{r}
d_correlations
```


# Correllate theoretical with empiricals, show noise ceiling

```{r}

d_correlations <- df_distances %>% 
  left_join(d_all_theories,
            by = c("itemA", "itemB")) %>% 
  group_by(subject_id) %>% 
  mutate(`1_Mind-Body-Action` = cor(emp_euclidean, `theor1_Mind-Body-Action`, method = "kendall"),
         `2_PerCog-ObjDir-Stimulus` = cor(emp_euclidean, `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`, method = "kendall"),
         `3_Physical-Ethereal` = cor(emp_euclidean, `theor3_Physical-Ethereal`, method = "kendall"),
         `4_cosine_similarity` = cor(emp_euclidean, `theor4_cosine_similarity`, method = "kendall")) %>% 
  select(-c("theor2_PerCog-OdActSpAct-BodyStimBodynoStim"))

    
d_correlations %>% 
  pivot_longer(cols = starts_with(c("1_","2_","3_","4_")),
               names_to = "correlation_type",
               values_to = "correlation_value") %>% 
  relocate(c("correlation_type", "correlation_value"), .after = "emp_euclidean") %>% 
  ggplot(aes(x = correlation_type, y = correlation_value)) +
  stat_summary(fun = "mean", geom = "bar", fill = "white", color = "black") +
  geom_point(aes(color = subject_id)) +
  theme(axis.text.x = element_text(angle = 10, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) +
  labs(x = "theoretical rdms", y = "Correlations",
       title = "  Correspondence between theoretical RDMS and subject-specific euclidean rdms") 

```

# correlating using upper only == to double check!

-  calculation was assuming that my data has diagonals, so my correlations were off. 
I was just correlating values from long form data and this included self correlations and thus inflated values. 

```{r}
# Function to extract upper triangular part as a vector
extract_upper_tri <- function(matrix) {
  matrix[upper.tri(matrix)]
}

# Compute Kendall's tau correlations using only the upper triangular part
#d_correlations <- 

d_correlations <- df_distances %>%
  left_join(d_all_theories, by = c("itemA", "itemB")) %>%
  group_by(subject_id) %>%
  mutate(
      `1_Mind-Body-Action` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor1_Mind-Body-Action`),   # Construct theoretical RDM
                          names_from = itemB, values_from = `theor1_Mind-Body-Action`) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
        
      `2_PerCog-ObjDir-Stimulus` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                            names_from = itemB, values_from = emp_euclidean) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`),     # Construct theoretical RDM
                            names_from = itemB, values_from = `theor2_PerCog-OdActSpAct-BodyStimBodynoStim`) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order]     
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
    
    `3_Physical-Ethereal` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor3_Physical-Ethereal`),     # Construct theoretical RDM
                          names_from = itemB, values_from = `theor3_Physical-Ethereal`) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        },
    
    `4_cosine_similarity` = {
              rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor4_cosine_similarity`),     # Construct theoretical RDM
                          names_from = itemB, values_from = `theor4_cosine_similarity`) %>%
        column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        }
    )

###-=====
# 
# d_correlations %>% 
#   pivot_longer(cols = starts_with(c("1_","2_","3_","4_")),
#                names_to = "correlation_type",
#                values_to = "correlation_value") %>% 
#   filter(subject_id == "subj_01") %>% 
#   arrange(desc(correlation_value)) %>% #cor with mind body action is .496. that can't be since noise ceiling is lower
#   ggplot(aes(x = correlation_type, y = correlation_value)) +
#   stat_summary(fun = "mean", geom = "bar", fill = "white", color = "black") +
#   geom_point((aes(color = subject_id))) +
#   theme(axis.text.x = element_text(angle = 10, hjust = 1, size = 12),
#         axis.text.y = element_text(size = 12),
#         axis.title.x = element_text(size = 14),
#         axis.title.y = element_text(size = 14)) +
#   labs(x = "theoretical rdms", y = "Correlations",
#        title = "  Correspondence between theoretical RDMS and subject-specific euclidean rdms") 

  
d_correlations %>% 
  filter(subject_id == "subj_01") %>% 
  # summarize(corelation = cor(emp_euclidean, `theor1_Mind-Body-Action`, method = "kendall"))
  select(c("subject_id", "emp_euclidean", `theor1_Mind-Body-Action`, `1_Mind-Body-Action`))

```

# rough sketchpad

```{r}
#subject 1 corr with MBA
df_distances %>%
  left_join(d_all_theories, by = c("itemA", "itemB")) %>%
  group_by(subject_id) %>%
  summarize(
      `1_Mind-Body-Action` = {
        rdm1 <- pivot_wider(pick(itemA, itemB, emp_euclidean),     #empirical RDM
                          names_from = itemB, values_from = emp_euclidean) %>%
                column_to_rownames("itemA") %>% as.matrix()
        
        rdm2 <- pivot_wider(pick(itemA, itemB, `theor1_Mind-Body-Action`),     # Construct theoretical RDM
                          names_from = itemB, values_from = `theor1_Mind-Body-Action`) %>%
               column_to_rownames("itemA") %>% as.matrix()
        
        rdm1 <- rdm1[desired_order, desired_order] 
        rdm2 <- rdm2[desired_order, desired_order]
        cor(rdm1[upper.tri(rdm1)], rdm2[upper.tri(rdm2)], method = "kendall")
        }) 
  

df_distances %>%
  left_join(d_all_theories, by = c("itemA", "itemB")) %>%
  filter(subject_id == "subj_01") %>% 
  select(itemA, itemB, `theor1_Mind-Body-Action`) %>% 
  pivot_wider(names_from = itemB, values_from = `theor1_Mind-Body-Action`) %>%
  column_to_rownames("itemA") %>% as.matrix()
        

  
#subj1 correlation with itself
# Subject's RDM
subject_rdm <- d_correlations %>%
  filter(subject_id == "subj_01") %>%
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
  column_to_rownames("itemA") %>%
  as.matrix()
subject_rdm <- subject_rdm[desired_order, desired_order]

#mind body rdm
mba_rdm <- d_correlations %>%
  filter(subject_id == "subj_01") %>%
  select(subject_id, itemA, itemB, `theor1_Mind-Body-Action`) %>% 
  pivot_wider(names_from = itemB, values_from = `theor1_Mind-Body-Action`) %>%
  column_to_rownames("itemA") %>%
  as.matrix()
mba_rdm <- mba_rdm[desired_order, desired_order]

# Grand average RDM
grand_avg_rdm <-  d_correlations %>%
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>%
  mutate(emp_euclidean = mean(emp_euclidean)) %>%
  ungroup() %>% 
  distinct(item_pairs, .keep_all = TRUE) %>% 
  select(-c(item_pairs, subject_id)) %>% 
  pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
  column_to_rownames("itemA") %>%
  as.matrix()
grand_avg_rdm <- grand_avg_rdm[desired_order, desired_order]

# Convert to numeric vectors
subject_upper <- as.numeric(subject_rdm[upper.tri(subject_rdm)])
grand_upper <- as.numeric(grand_avg_rdm[upper.tri(grand_avg_rdm)])
mba_upper <- as.numeric(mba_rdm[upper.tri(mba_rdm)])

# Compute correlations
upper_bound <- cor(subject_upper, other_upper, method = "kendall")
lower_bound <- cor(subject_upper, grand_upper, method = "kendall") #but this is showing me a higher value than upper
mba_corr <- cor(subject_upper, mba_upper, method = "kendall") #but this is showing me a higher value than upper


```

noise ceiling for correlations

```{r}

desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")
df_distances
d_all_theories
d_correlations


# Compute noise ceiling for all subjects
subjects <- unique(d_correlations$subject_id)

noise_ceiling <- calculate_noise_ceiling(d_correlations, subjects, desired_order)

upper_bound <- numeric(length(subjects))
lower_bound <- numeric(length(subjects))

#subject's RDM
subject_rdm <- d_correlations %>%
  filter(subject_id == "subj_01") %>%
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
  column_to_rownames("itemA") %>%
  as.matrix()
subject_rdm <- subject_rdm[desired_order, desired_order]

#grand average RDM
grand_avg_rdm <-  d_correlations %>%
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>%
  mutate(emp_euclidean = mean(emp_euclidean)) %>%
  ungroup() %>% 
  distinct(item_pairs, .keep_all = TRUE) %>% 
  select(-c(item_pairs, subject_id)) %>% 
  pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
  column_to_rownames("itemA") %>%
  as.matrix()
grand_avg_rdm <- grand_avg_rdm[desired_order, desired_order]

#other subjects' average RDM
other_avg_rdm <- d_correlations %>%
  select(subject_id, itemA, itemB, emp_euclidean) %>% 
  filter(subject_id != "subj_01") %>%
  unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
  group_by(item_pairs) %>%
  mutate(emp_euclidean = mean(emp_euclidean)) %>%
  ungroup() %>% 
  distinct(item_pairs, .keep_all = TRUE) %>% 
  select(-c(item_pairs, subject_id)) %>% 
  pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
  column_to_rownames("itemA") %>%
  as.matrix()
other_avg_rdm <- other_avg_rdm[desired_order, desired_order]
    
# Convert to numeric vectors
subject_upper <- as.numeric(subject_rdm[upper.tri(subject_rdm)])
other_upper <- as.numeric(other_avg_rdm[upper.tri(other_avg_rdm)])
grand_upper <- as.numeric(grand_avg_rdm[upper.tri(grand_avg_rdm)])

# Compute correlations
lower_bound <- cor(subject_upper, other_upper, method = "kendall")
upper_bound <- cor(subject_upper, grand_upper, method = "kendall") #but this is showing me a higher value than upper


df_matrix <- d_correlations %>% 
    select(subject_id, itemA, itemB, emp_euclidean) %>% 
    filter(subject_id == "subj_01") %>%
    pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
    column_to_rownames("itemA") %>%
    as.matrix()
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0  # Set diagonal to 0
  
  # Ensure df_matrix is numeric
  df_matrix <- apply(df_matrix, 2, as.numeric)
  
  # Plot RDM
  corrplot(
    as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = paste("RDM -", subject),
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
  
#==
plot_rdm_with_ceiling <- function(data, subject, desired_order, noise_ceiling) {
  # Prepare subject's RDM
  df_matrix <- data %>%
    filter(subject_id == subject) %>%
    pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
    column_to_rownames("itemA") %>%
    as.matrix()
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0  # Set diagonal to 0
  
  # Ensure df_matrix is numeric
  df_matrix <- apply(df_matrix, 2, as.numeric)
  
  # Plot RDM
  corrplot(
    as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = paste("RDM -", subject),
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
  
  # Add noise ceiling annotations
  if (!is.null(noise_ceiling)) {
    mtext(
      text = paste0("Noise Ceiling:\nUpper: ", round(as.numeric(noise_ceiling$upper_bound), 3),
                    "\nLower: ", round(as.numeric(noise_ceiling$lower_bound), 3)),
      side = 3, line = -2, adj = 1, cex = 0.8
    )
  }
}

noise_ceiling <- calculate_noise_ceiling(df_distances, subjects, desired_order)



```


```{r}
calculate_noise_ceiling <- function(data, subjects, desired_order, metric = "kendall") {
  # Initialize storage for correlations
  upper_bound <- numeric(length(subjects))
  lower_bound <- numeric(length(subjects))
  
  # Loop through each subject
  for (i in seq_along(subjects)) {
    subject <- subjects[i]
    
    # Subject's RDM
    subject_rdm <- data %>%
        filter(subject_id == subject) %>%
        select(subject_id, itemA, itemB, emp_euclidean) %>% 
        pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    subject_rdm <- subject_rdm[desired_order, desired_order]
    
    # Grand average RDM
    grand_avg_rdm <- data %>%
        select(subject_id, itemA, itemB, emp_euclidean) %>% 
        unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
        group_by(item_pairs) %>%
        mutate(emp_euclidean = mean(emp_euclidean)) %>%
        ungroup() %>% 
        distinct(item_pairs, .keep_all = TRUE) %>% 
        select(-c(item_pairs, subject_id)) %>% 
        pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
        column_to_rownames("itemA") %>%
        as.matrix()
    grand_avg_rdm <- grand_avg_rdm[desired_order, desired_order]
    
    # Other subjects' average RDM
    other_avg_rdm <- data %>%
      select(subject_id, itemA, itemB, emp_euclidean) %>% 
      filter(subject_id != "subj_01") %>%
      unite("item_pairs", itemA, itemB, sep = "_", remove = FALSE) %>% 
      group_by(item_pairs) %>%
      mutate(emp_euclidean = mean(emp_euclidean)) %>%
      ungroup() %>% 
      distinct(item_pairs, .keep_all = TRUE) %>% 
      select(-c(item_pairs, subject_id)) %>% 
      pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
      column_to_rownames("itemA") %>%
      as.matrix()
    other_avg_rdm <- other_avg_rdm[desired_order, desired_order]
    
    # Convert to numeric vectors
    subject_upper <- as.numeric(subject_rdm[upper.tri(subject_rdm)])
    other_upper <- as.numeric(other_avg_rdm[upper.tri(other_avg_rdm)])
    grand_upper <- as.numeric(grand_avg_rdm[upper.tri(grand_avg_rdm)])
    
    # Compute correlations
    upper_bound[i] <- cor(subject_upper, other_upper, method = metric)
    lower_bound[i] <- cor(subject_upper, grand_upper, method = metric)
  }
  
  # Return average bounds
  list(
    upper_bound = mean(upper_bound, na.rm = TRUE),
    lower_bound = mean(lower_bound, na.rm = TRUE)
  )
}

plot_rdm_with_ceiling <- function(data, subject, desired_order, noise_ceiling) {
  # Prepare subject's RDM
  df_matrix <- data %>%
    select(subject_id, itemA, itemB, emp_euclidean) %>% 
    filter(subject_id == subject) %>%
    pivot_wider(names_from = itemB, values_from = emp_euclidean) %>%
    column_to_rownames("itemA") %>%
    as.matrix()
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0  # Set diagonal to 0
  
  # Ensure df_matrix is numeric
  df_matrix <- apply(df_matrix, 2, as.numeric)
  
  # Plot RDM
  corrplot(
    as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = paste("RDM -", subject),
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
  
  # Add noise ceiling annotations
  if (!is.null(noise_ceiling)) {
    mtext(
      text = paste0("Noise Ceiling:\nUpper: ", round(as.numeric(noise_ceiling$upper_bound), 3),
                    "\nLower: ", round(as.numeric(noise_ceiling$lower_bound), 3)),
      side = 3, line = -2, adj = 1, cex = 0.8
    )
  }
}

# Compute noise ceiling for all subjects
subjects <- unique(d_correlations$subject_id)
noise_ceiling <- calculate_noise_ceiling(d_correlations, subjects, desired_order)

# Generate plots
for (subject in subjects) {
  plot_rdm_with_ceiling(df_distances, subject, desired_order, noise_ceiling)
}

##=================Plotting correlations
d_correlations %>%
  pivot_longer(cols = starts_with(c("1_","2_","3_","4_")),
               names_to = "correlation_type",
               values_to = "correlation_value") %>%
  ggplot(aes(x = correlation_type, y = correlation_value)) +
  stat_summary(fun = "mean", geom = "bar", fill = "white", color = "black") +
  geom_point(aes(color = subject_id)) +
  geom_hline(yintercept = noise_ceiling$upper_bound, color = "blue", linetype = "dashed", size = 1) +
  geom_hline(yintercept = noise_ceiling$lower_bound, color = "red", linetype = "dashed", size = 1) +
  theme(axis.text.x = element_text(angle = 10, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14)) +
  labs(x = "Theoretical RDMs", y = "Correlations",
       title = "Correspondence between theoretical RDMs and subject-specific Euclidean RDMs",
       subtitle = paste0("Noise Ceiling:\nUpper: ", round(noise_ceiling$upper_bound, 3),
                         " | Lower: ", round(noise_ceiling$lower_bound, 3)))

```



# check feedback

```{r}
df_feedback <- d %>% 
  select(feedback) %>% 
  drop_na(feedback)

cat(df_feedback$feedback, sep = "\n\n")
  
```

