---
title: "study 6"
output: html_document
date: "2024-10-28"
---

# ReadMe

This is the analysis for study 6 of the MBC project, alias the counterfactual study.  
The question: 

# Set options

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(
               lme4, 
               here, 
               patchwork, 
               cowplot, 
               viridis, 
               effects, 
               gt,
               webshot2,
               knitr,
               performance,
               tidyverse)
sessionInfo()

```

# Global variables

```{r}
base_path <- here("code", "analysis", "pilot")
freesort_and_causal_distances_df_path <- paste0(base_path, "/data-explanation_1/freesort_minus_causals.csv")

```


# Study 3 - Causality for intervention and inference

## Prepare data

### read data

```{r, warnings = FALSE}
data_path_cf <- paste0(base_path, "/data-study6/pilot1")

desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

file_paths <- list.files(path = data_path_cf, pattern = "^2025-", full.names = TRUE)

d <- imap_dfr(file_paths, ~ {
  read_csv(.x, col_types = cols(response = col_character())) %>%
    mutate(
      filename = basename(.x),
      subject_id = paste0("subj_", str_pad(.y, 2, pad = "0"))
    )}) %>%
  select(-c(mturk_participant_id, mturk_assignment_id, mturk_project_id)) %>%
  relocate(subject_id, .before = "success")

d %>% 
  filter(condition == "causal")

```

### wrangle data

initial: 30 trials per subject * 48 subjects = 1440 rows.
dat: (30 trials / 2 trials per row) * 48 subjects = 720 rows.

(after filtering to only relevant trials)
initial: 15 trials per subject * 2 subjects = 30 rows

TODO: rename itemB to target item.

```{r}
subjects_to_exclude <- d %>%
  filter(grepl("attention_check", condition)) %>% 
  mutate(attn_correct = correct_response_attn == selected_choice) %>% 
  filter(attn_correct == FALSE) %>%
  pull(subject_id) %>% 
  unique()


dat <- d %>% 
  filter(!subject_id %in% subjects_to_exclude) %>% #exclude 1 subject who failed attn check
  filter(condition %in% c("causal", "similar")) %>%
  mutate(response = as.factor(response)) %>% 
  mutate(condition = as.factor(condition)) %>% 
  rowwise() %>%
  mutate(response = case_when(
    grepl(causal_itemA, selected_choice) ~ "causal",
    grepl(similarity_itemA, selected_choice) ~ "similar",
    TRUE ~ "other"
  )) %>% 
  ungroup() %>% 
  select(subject_id, condition, itemB, causal_itemA, similarity_itemA, response) %>% 
  mutate(causal_itemA = factor(causal_itemA, levels = desired_order),
         similarity_itemA = factor(similarity_itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(domain_itemB = case_when(itemB == "hear something" ~ "mind", #labels for first item of pair
                                 itemB == "choose what to do" ~ "mind",
                                 itemB == "remember something" ~ "mind",
                                 itemB == "think about something" ~ "mind",
                                 itemB == "see something" ~ "mind",
                                 itemB == "reach for something" ~ "action",
                                 itemB == "sit down" ~ "action", 
                                 itemB == "jump up and down" ~ "action",
                                 itemB == "kick something" ~ "action",
                                 itemB == "take a walk" ~ "action",
                                 itemB == "get tired" ~ "bio", 
                                 itemB == "become hungry" ~ "bio",
                                 itemB == "feel scared"~ "bio",
                                 itemB == "experience pain"~ "bio",
                                 itemB == "get sick"~ "bio"),
         domain_causalitemA = case_when(causal_itemA == "hear something" ~ "mind", 
                                 causal_itemA == "choose what to do" ~ "mind",
                                 causal_itemA == "remember something" ~ "mind",
                                 causal_itemA == "think about something" ~ "mind",
                                 causal_itemA == "see something" ~ "mind",
                                 causal_itemA == "reach for something" ~ "action",
                                 causal_itemA == "sit down" ~ "action", 
                                 causal_itemA == "jump up and down" ~ "action",
                                 causal_itemA == "kick something" ~ "action",
                                 causal_itemA == "take a walk" ~ "action",
                                 causal_itemA == "get tired" ~ "bio", 
                                 causal_itemA == "become hungry" ~ "bio",
                                 causal_itemA == "feel scared"~ "bio",
                                 causal_itemA == "experience pain"~ "bio",
                                 causal_itemA == "get sick"~ "bio"),
          domain_similarityitemA = case_when(similarity_itemA == "hear something" ~ "mind", 
                                 similarity_itemA == "choose what to do" ~ "mind",
                                 similarity_itemA == "remember something" ~ "mind",
                                 similarity_itemA == "think about something" ~ "mind",
                                 similarity_itemA == "see something" ~ "mind",
                                 similarity_itemA == "reach for something" ~ "action",
                                 similarity_itemA == "sit down" ~ "action", 
                                 similarity_itemA == "jump up and down" ~ "action",
                                 similarity_itemA == "kick something" ~ "action",
                                 similarity_itemA == "take a walk" ~ "action",
                                 similarity_itemA == "get tired" ~ "bio", 
                                 similarity_itemA == "become hungry" ~ "bio",
                                 similarity_itemA == "feel scared"~ "bio",
                                 similarity_itemA == "experience pain"~ "bio",
                                 similarity_itemA == "get sick"~ "bio")) 




```


```{r}
dat_summarized <- dat %>%
  mutate(response = fct_relevel(response, c("similar", "causal"))) %>% 
  group_by(condition) %>%
  mutate(total = n()) %>% 
  group_by(condition, response) %>%
  summarize(
    count = n(),  
    total = first(total), 
    mean_proportion = count / total, 
    se = sqrt(mean_proportion * (1 - mean_proportion) / total),  
    .groups = "drop")

#stacked
(fig_result_summary <- ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_manual(values = c("grey80", "#000000")) + #or whatever you'd like!
  geom_errorbar(data = filter(dat_summarized, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  theme_cowplot() +
  labs(y = "Proportion"))

#visualize bar heights for writeup
(fig_result_summary <- ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_manual(values = c("grey80", "#000000")) + # Customize bar colors
  geom_errorbar(data = filter(dat_summarized, response == "causal"), 
                aes(ymax = mean_proportion + se, ymin = mean_proportion - se), 
                stat = "identity", position = "identity", width = 0.3) +
  geom_text(aes(label = round(mean_proportion, 2), 
                y = mean_proportion + 0.02), # Adjust the y position slightly above the bar
            position = position_dodge(width = 0.9), 
            vjust = 0,
            color = "white") + # Adjust vertical alignment
  theme_cowplot())


#fig_result_summary <- 
ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_point() +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_manual(values = c("grey80", "#000000")) + #or whatever you'd like!
  geom_errorbar(data = filter(dat_summarized, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  theme_cowplot() +
  labs(y = "Proportion")

 dat %>%
  mutate(response = fct_relevel(response, c("similar", "causal"))) %>% 
  group_by(condition) %>%
  mutate(total = n()) %>% 
  group_by(condition, response) %>%
  summarize(
    count = n(),  
    total = first(total))
 
 
```




