---
title: "study 2"
output: html_document
date: "2024-10-28"
---

# ReadMe

This is the analysis for study 3 of the MBC project, alias the counterfactual study.  
The study had N = XX, and subjects saw two conditions: Intervention and Similarity, with 15 trials per condition.
The question: 

# Set options

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(tidyverse, 
               lme4, 
               here, 
               patchwork, 
               cowplot, 
               viridis, 
               effects, 
               gt,
               webshot2,
               knitr,
               performance)
sessionInfo()

```

# Read data


```{r, warnings = FALSE}
base_path <- here("code", "analysis", "pilot")
data_path <- paste0(base_path, "/data-counterfactual_1/pilot")

desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

d <- read.csv(paste0(data_path, "/2025-03-25_13-12-24_undefined.csv"))

#do this in a separate pre-processing pipeline
file_paths <- list.files(path = data_path, pattern = "^2025-", full.names = TRUE)
d <- map_df(file_paths, ~ read_csv(.x) %>% mutate(filename = basename(.x))) %>%
     mutate(subject_id = paste0("subj_", cur_group_id())) %>%
     select(-c(mturk_participant_id, mturk_assignment_id, mturk_project_id)) %>%
     relocate(subject_id, .before = "success")

```


# Wrangle data

```{r}
 subjects_to_exclude <- d %>%
  filter(grepl("attention_check", condition)) %>% 
  filter(selected_choice == "no") %>% 
  pull(subject_id) %>% 
  unique()

dat <- d %>% 
  # filter(!subject_id %in% subjects_to_exclude) %>% 
  filter(subject_id %in% subjects_to_exclude) %>% #exclude 1 subject who failed attn check
  filter(condition %in% c("intervention", "similarity")) %>%
  mutate(response = as.factor(response)) %>% 
  mutate(condition = as.factor(condition)) %>% 
  rowwise() %>%
  mutate(response = case_when(
    grepl(causal_itemA, selected_choice) ~ "causal",
    grepl(similarity_itemA, selected_choice) ~ "similar",
    TRUE ~ "other"
  )) %>% 
  mutate(condition = case_when(
    condition == "similarity" ~ "inference",
    TRUE ~ condition
  )) %>% 
  ungroup() %>% 
  select(subject_id, condition, itemB, causal_itemA, similarity_itemA, response) %>% 
  mutate(causal_itemA = factor(causal_itemA, levels = desired_order),
         similarity_itemA = factor(similarity_itemA, levels = desired_order),
         itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(domain_itemB = case_when(itemB == "hear something" ~ "mind", #labels for first item of pair
                                 itemB == "choose what to do" ~ "mind",
                                 itemB == "remember something" ~ "mind",
                                 itemB == "think about something" ~ "mind",
                                 itemB == "see something" ~ "mind",
                                 itemB == "reach for something" ~ "action",
                                 itemB == "sit down" ~ "action", 
                                 itemB == "jump up and down" ~ "action",
                                 itemB == "kick something" ~ "action",
                                 itemB == "take a walk" ~ "action",
                                 itemB == "get tired" ~ "bio", 
                                 itemB == "become hungry" ~ "bio",
                                 itemB == "feel scared"~ "bio",
                                 itemB == "experience pain"~ "bio",
                                 itemB == "get sick"~ "bio"),
         domain_causalitemA = case_when(causal_itemA == "hear something" ~ "mind", 
                                 causal_itemA == "choose what to do" ~ "mind",
                                 causal_itemA == "remember something" ~ "mind",
                                 causal_itemA == "think about something" ~ "mind",
                                 causal_itemA == "see something" ~ "mind",
                                 causal_itemA == "reach for something" ~ "action",
                                 causal_itemA == "sit down" ~ "action", 
                                 causal_itemA == "jump up and down" ~ "action",
                                 causal_itemA == "kick something" ~ "action",
                                 causal_itemA == "take a walk" ~ "action",
                                 causal_itemA == "get tired" ~ "bio", 
                                 causal_itemA == "become hungry" ~ "bio",
                                 causal_itemA == "feel scared"~ "bio",
                                 causal_itemA == "experience pain"~ "bio",
                                 causal_itemA == "get sick"~ "bio"),
          domain_similarityitemA = case_when(similarity_itemA == "hear something" ~ "mind", 
                                 similarity_itemA == "choose what to do" ~ "mind",
                                 similarity_itemA == "remember something" ~ "mind",
                                 similarity_itemA == "think about something" ~ "mind",
                                 similarity_itemA == "see something" ~ "mind",
                                 similarity_itemA == "reach for something" ~ "action",
                                 similarity_itemA == "sit down" ~ "action", 
                                 similarity_itemA == "jump up and down" ~ "action",
                                 similarity_itemA == "kick something" ~ "action",
                                 similarity_itemA == "take a walk" ~ "action",
                                 similarity_itemA == "get tired" ~ "bio", 
                                 similarity_itemA == "become hungry" ~ "bio",
                                 similarity_itemA == "feel scared"~ "bio",
                                 similarity_itemA == "experience pain"~ "bio",
                                 similarity_itemA == "get sick"~ "bio")) 


d %>% 
  count(condition)

```

# Plot

## table of choices

TODO add mean causal and sim distances between A and B? --> done
```{r}
# design_pairs <- dat %>%
#   select(itemB, causal_itemA, similarity_itemA) %>%
#   unique() %>%
#   relocate(itemB, .after = last_col()) %>%
#   rename(target_item = itemB, causal_choice = causal_itemA, similar_choice = similarity_itemA)

design_pairs <- dat %>% #start with raw data containing items
  select(itemB, causal_itemA, similarity_itemA) %>%
  unique() %>%
  left_join(
    read.csv(paste0(snapshots_input_path, "/diffs.csv")) %>% #grab causal item distances
      mutate(
        itemB = factor(itemB, levels = desired_order),
        itemA = factor(itemA, levels = desired_order)
      ) %>%
      select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean),
    by = c("itemB", "causal_itemA" = "itemA")
  ) %>%
  left_join(
    read.csv(paste0(snapshots_input_path, "/diffs.csv")) %>% #grab similar item distances
      mutate(
        itemB = factor(itemB, levels = desired_order),
        itemA = factor(itemA, levels = desired_order)
      ) %>%
      select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean),
    by = c("itemB", "similarity_itemA" = "itemA"),
    suffix = c("_causal", "_similar")
  ) %>%
  distinct(itemB, .keep_all = TRUE) %>%
  mutate(
    mean_causal_dissimilarity_causal = round(mean_causal_dissimilarity_causal, 2),
    mean_emp_euclidean_causal = round(mean_emp_euclidean_causal, 2),
    mean_causal_dissimilarity_similar = round(mean_causal_dissimilarity_similar, 2),
    mean_emp_euclidean_similar = round(mean_emp_euclidean_similar, 2)
  ) %>%
  select(
    target_item = itemB,
    causal_choice = causal_itemA,
    causal_distance_causal_choice = mean_causal_dissimilarity_causal,
    freesort_distance_causal_choice = mean_emp_euclidean_causal,
    similar_choice = similarity_itemA,
    causal_distance_similar_choice = mean_causal_dissimilarity_similar,
    freesort_distance_similar_choice = mean_emp_euclidean_similar
  ) %>%
  arrange(target_item)

```

## table plotting with gt

```{r}
#table with line breaks
pairs_table <- design_pairs %>%
  mutate(
    causal_choice = paste0(causal_choice, "<br>(", causal_distance_causal_choice, ", ", freesort_distance_causal_choice, ")"),
    similar_choice = paste0(similar_choice, "<br>(", causal_distance_similar_choice, ", ", freesort_distance_similar_choice, ")"),
    group = case_when(
      row_number() %in% 1:5 ~ "mind",
      row_number() %in% 6:10 ~ "action",
      row_number() %in% 11:15 ~ "body"
    )
  ) %>%
  select(target_item, causal_choice, similar_choice, group)

# Generate GT table with markdown rendering
df_table <- pairs_table %>%
  gt() %>%
  tab_header(
    title = "Target and Choice Items",
    subtitle = "Distances Between Target, Causal, and Similar Choices"
  ) %>%
  cols_label(
    target_item = "Target Item",
    causal_choice = md("Causal Choice<br>(Causal D., Freesort D.)"),
    similar_choice = md("Similar Choice<br>(Causal D., Freesort D.)")
  ) %>%
  fmt_markdown(columns = c(causal_choice, similar_choice)) %>%  # Interpret markdown for line breaks
  tab_style(
    style = cell_fill(color = "#f7bf2a"),  # Highlight for 'mind'
    locations = cells_body(rows = group == "mind")
  ) %>%
  tab_style(
    style = cell_fill(color = "#69ad44"),  # Highlight for 'action'
    locations = cells_body(rows = group == "action")
  ) %>%
  tab_style(
    style = cell_fill(color = "#f36b2d"),  # Highlight for 'body'
    locations = cells_body(rows = group == "body")
  ) %>%
  cols_hide(columns = group) %>%   # Optionally hide the helper column
  tab_options(
    table.font.names = "Baskerville"
  )

# Save
# gtsave(df_table, "pairs-table.html")  # Save as HTML for viewing with color
gtsave(df_table, "pairs-table.png")   # Save as PNG for use in presentations or documents
```

## table without line breaks

```{r}
#==============omitting line breaks
pairs_table <- design_pairs %>%
  mutate(
    causal_choice = paste0(causal_choice, " (", causal_distance_causal_choice, ", ", freesort_distance_causal_choice, ")"),
    similar_choice = paste0(similar_choice, " (", causal_distance_similar_choice, ", ", freesort_distance_similar_choice, ")"),
    group = case_when(
      row_number() %in% 1:5 ~ "mind",
      row_number() %in% 6:10 ~ "action",
      row_number() %in% 11:15 ~ "body"
    )
  ) %>%
  select(target_item, causal_choice, similar_choice, group)

# Generate GT table with markdown rendering
df_table <- pairs_table %>%
  gt() %>%
  tab_header(
    title = "Target and Choice Items" #,
    # subtitle = "Distances Between Target, Causal, and Similar Choices"
  ) %>%
  cols_label(
    target_item = "Target Item",
    causal_choice = md("Causal Choice<br>(Causal D., Freesort D.)"),
    similar_choice = md("Similar Choice<br>(Causal D., Freesort D.)")
  ) %>%
  fmt_markdown(columns = c(causal_choice, similar_choice)) %>%  # Interpret markdown for line breaks
  tab_style(
    style = cell_fill(color = "#f7bf2a"),  # Highlight for 'mind'
    locations = cells_body(rows = group == "mind")
  ) %>%
  tab_style(
    style = cell_fill(color = "#69ad44"),  # Highlight for 'action'
    locations = cells_body(rows = group == "action")
  ) %>%
  tab_style(
    style = cell_fill(color = "#f36b2d"),  # Highlight for 'body'
    locations = cells_body(rows = group == "body")
  ) %>%
  cols_hide(columns = group) %>%   # Optionally hide the helper column
  tab_options(
    table.font.names = "Baskerville"
  )

gtsave(df_table, "pairs-table-2.png")   # Save as PNG for use in presentations or documents

```



## overall barplot

TODO add error bars to stacked barplot? --> added
TODO decide what to call conditions and responses - conditions = intervention vs inference? and response = similar vs causal? --> yup, changed
TODO make causal responses at bottom since that's the unique contribution? --> done
```{r}
dat_summarized <- dat %>%
  mutate(response = fct_relevel(response, c("similar", "causal"))) %>% 
  group_by(condition) %>%
  mutate(total = n()) %>% 
  group_by(condition, response) %>%
  summarize(
    count = n(),  
    total = first(total), 
    mean_proportion = count / total, 
    se = sqrt(mean_proportion * (1 - mean_proportion) / total),  
    .groups = "drop")

#stacked
(fig_result_summary <- ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_manual(values = c("grey80", "#000000")) + #or whatever you'd like!
  geom_errorbar(data = filter(dat_summarized, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  theme_cowplot() +
  labs(y = "Proportion"))

#visualize bar heights for writeup
(fig_result_summary <- ggplot(dat_summarized, aes(x = condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity", colour = "black") +
  scale_fill_manual(values = c("grey80", "#000000")) + # Customize bar colors
  geom_errorbar(data = filter(dat_summarized, response == "causal"), 
                aes(ymax = mean_proportion + se, ymin = mean_proportion - se), 
                stat = "identity", position = "identity", width = 0.3) +
  geom_text(aes(label = round(mean_proportion, 2), 
                y = mean_proportion + 0.02), # Adjust the y position slightly above the bar
            position = position_dodge(width = 0.9), 
            vjust = 0,
            color = "white") + # Adjust vertical alignment
  theme_cowplot())

```

## modelling barplot

```{r}
dat_prepped <- dat %>%
  mutate(
    response_binary = ifelse(response == "causal", 1, 0), # Binary outcome (1 = causal)
    condition = as.factor(condition) # Ensure condition is a factor
  )

model <- glmer(response_binary ~ condition + (1 | subject_id), 
               data = dat_prepped, 
               family = binomial)

# Summary of the model
summary(model)

```


## item level, stacked

```{r, fig.width = 12}
dat_summarized_by_item <- dat %>% 
  mutate(response = fct_relevel(response, c("similar", "causal"))) %>% 
  group_by(condition, itemB) %>%
  mutate(total = n()) %>%  # Total responses per condition
  group_by(condition, itemB, response) %>%
  mutate(
    count = n(),  # Count responses for each response type
    total = first(total),  # Total responses for the condition
    mean_proportion = count / total,  # Proportion of response within condition
    se = sqrt(mean_proportion * (1 - mean_proportion) / total)) %>% 
  distinct(condition, itemB, .keep_all = TRUE) %>% 
  ungroup()

#set x axis colors
itemB_colors <- dat_summarized_by_item %>%
  distinct(itemB, domain_itemB) %>%
  mutate(color = case_when(
    domain_itemB == "bio" ~ "darkgreen",
    domain_itemB == "mind" ~ "blue",
    domain_itemB == "action" ~ "red"
  )) %>%
  pull(color, name = itemB)

p <- dat_summarized_by_item %>% 
  ggplot(aes(x =itemB, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity") +
  geom_bar(stat = "identity", colour = "black") +
  geom_errorbar(data = filter(dat_summarized_by_item, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  scale_fill_manual(values = c("grey80", "#000000")) + #or whatever you'd like!
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(dat_summarized_by_item$itemB)])) +
  facet_wrap(~condition) 

p

p_alt <- dat_summarized_by_item %>% 
  ggplot(aes(x =condition, y = mean_proportion, fill = response)) +
  geom_bar(stat = "identity") +
  geom_bar(stat = "identity", colour = "black") +
  geom_errorbar(data = filter(dat_summarized_by_item, response == "causal"), aes(ymax = mean_proportion + se,
                                                                         ymin = mean_proportion - se),
                stat = "identity", position = "identity", width = 0.3) +
  scale_fill_manual(values = c("grey80", "#000000")) + #or whatever you'd like!
  theme_cowplot() +
  # theme(axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(dat_summarized_by_item$itemB)])) +
  facet_wrap(~itemB, nrow = 3) 

p_alt

ggsave(
  filename = "figures/study2/p_study2_item-level_stacked.pdf",
  plot = p,          
  device = "pdf",           
  #path = "path/to/directory", #Optional: specify a directory
  width = 12,               
  height = 6,             
  units = "in"              
)

```

# Modelling

## effect of condition

Todo: rename mean_emp_euclidean and mean_causal_dissimilarity to causal and freesort distance in all scripts!

```{r}
d_model <- dat %>% 
  mutate(response = as.factor(response)) %>% 
  mutate(subject_id = as.factor(subject_id)) %>% 
  mutate(responses = ifelse(response == "causal", 1, 0)) %>% 
  mutate(condition = as.factor(condition)) %>%
  unite(col = "trial_ID", c(itemB, similarity_itemA, causal_itemA), sep = "_", remove = FALSE)
  
#model
overall_model <- glmer(responses ~ condition + (1|subject_id), data = d_model, family = binomial(link = 'logit'))
summary(overall_model)
plot(allEffects(overall_model))
check_model(overall_model) # TODO figure out binned resisual issue?

```

## effect of causal and freesort distance in causal condition

TODO convert CIs into probs too?
TODO give models unique names

```{r}
d_freesort_and_causal_distances <- read.csv(paste0(snapshots_input_path, "/diffs.csv")) %>% 
  mutate(itemB = factor(itemB, levels = desired_order)) %>% 
  mutate(itemA = factor(itemA, levels = desired_order)) 

causal_condition <- d_model %>% 
  # select(-similarity_itemA) %>% # why - let's study both?
    filter(condition == "intervention") %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean, euclidean_minus_causal) %>% 
                rename("causal_itemA" = itemA,
                       "causal_itemA_causal_distance" = mean_causal_dissimilarity,
                       "causal_itemA_freesort_distance" = mean_emp_euclidean,
                       "causal_itemA_freesort_minus_causal" = euclidean_minus_causal),
              by = c("itemB", "causal_itemA")) %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean, euclidean_minus_causal) %>% 
                rename("similarity_itemA" = itemA,
                       "similarity_itemA_causal_distance" = mean_causal_dissimilarity,
                       "similarity_itemA_freesort_distance" = mean_emp_euclidean, 
                       "similar_itemA_freesort_minus_causal" = euclidean_minus_causal),
              by = c("itemB", "similarity_itemA"))

  
similarity_condition <- d_model %>% 
  # select(-causal_itemA) %>% 
  filter(condition == "similarity") %>% 
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean, euclidean_minus_causal) %>% 
                rename("causal_itemA" = itemA,
                       "causal_itemA_causal_distance" = mean_causal_dissimilarity,
                       "causal_itemA_freesort_distance" = mean_emp_euclidean,
                       "causal_itemA_freesort_minus_causal" = euclidean_minus_causal),
              by = c("itemB", "causal_itemA")) %>%
    left_join(d_freesort_and_causal_distances %>% 
                select(itemA, itemB, mean_causal_dissimilarity, mean_emp_euclidean, euclidean_minus_causal) %>% 
                rename("similarity_itemA" = itemA,
                       "similarity_itemA_causal_distance" = mean_causal_dissimilarity,
                       "similarity_itemA_freesort_distance" = mean_emp_euclidean, 
                       "similar_itemA_freesort_minus_causal" = euclidean_minus_causal),
              by = c("itemB", "similarity_itemA"))


#Modelling both predictors 
both_predictors_model <- glmer(responses ~ causal_itemA_freesort_minus_causal + similar_itemA_freesort_minus_causal + (1|subject_id), data = causal_condition, family = binomial(link = "logit"))
summary(both_predictors_model)
plot(allEffects(both_predictors_model))
check_model(both_predictors_model)

fixed_effects <- fixef(both_predictors_model) #log odds
odds_ratios <- exp(fixed_effects)  #convert log-odds to odds ratios
probabilities <- 1 / (1 + exp(-fixed_effects)) #convert log odds to probability

cat("\nBoth predictors model [intercept] [causal_distance] [freesort_distance]:\n ")
cat("Fixed effects (log odds): ", fixed_effects, "\n")
cat("Fixed effects (odds ratios): ", odds_ratios, "\n")
cat("Fixed effects (probabilities, causal_distance = 0):", probabilities, "\n")
#nb: fixed effects are intercept and slope. slope is change log odds of outcome for a unit increase in predictor

```


## plotting effects combined graph

```{r, fig.width = 8, fig.height = 8}
effects_data <- allEffects(both_predictors_model)

causal_effect <- as.data.frame(effects_data$`causal_itemA_freesort_minus_causal`)
similar_effect <- as.data.frame(effects_data$`similar_itemA_freesort_minus_causal`)

causal_effect$Predictor <- "Intervention"
similar_effect$Predictor <- "Inference"

colnames(causal_effect)[1] <- "PredictorValue"
colnames(similar_effect)[1] <- "PredictorValue"

combined_effects <- rbind(causal_effect, similar_effect)

base_plot <- ggplot(combined_effects, aes(x = PredictorValue, y = fit, color = Predictor)) +
  geom_line() +
  geom_hline(yintercept = .5, linetype = "dashed") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = Predictor), alpha = 0.2) +
  labs(
    title = "Probability of Making A Causal Choice",
    x = "Choice Item Distance to Target Item",
    y = "Probability",
    color = "Condition",
    fill = "Condition"
  ) +
  theme_cowplot() +
  # scale_x_continuous(
  #   breaks = c(min(combined_effects$PredictorValue), max(combined_effects$PredictorValue)),
  #   labels = c("More similar, \nless causal", "More causal, \nless similar")
  # ) +
  theme(
    plot.title = element_text(hjust = 0.5),  
    axis.text.x = element_text(size = 16),  
    axis.text.y = element_text(size = 16),  
    axis.title.y = element_text(size = 16),  
    axis.title.x = element_text(size = 16),  
    # legend.position = "bottom", 
    legend.title = element_text(size = 16), 
    legend.text = element_text(size = 16),
    plot.margin = margin(20, 20, 50, 20)   
  )


# Add annotations below the x-axis using cowplot's ggdraw
ggdraw(base_plot) +
  draw_label(
    label = "More similar,\nless causal",
    x = 0.1, y = 0.02, hjust = 0, vjust = 0, size = 16
  ) +
  draw_label(
    label = " More causal,\nless similar",
    x = 0.8, y = 0.02, hjust = 1, vjust = 0, size = 16
  )


```



## effect of causal and freesort distance in similarity condition

TODO check the sign of the coefficients / order of DV levels here?
```{r}
# Modelling predictors jointly
both_predictors_model <- glmer(responses ~ causal_itemA_freesort_minus_causal + similar_itemA_freesort_minus_causal + (1|subject_id), data = similarity_condition, family = binomial(link = "logit")) # TODO figure out failure to converge here
summary(both_predictors_model)
plot(allEffects(both_predictors_model))
check_model(both_predictors_model)

fixed_effects <- fixef(both_predictors_model) #log odds
odds_ratios <- exp(fixed_effects)  #convert log-odds to odds ratios
probabilities <- 1 / (1 + exp(-fixed_effects)) #convert log odds to probability

cat("\nBoth predictors model [intercept] [causal_distance] [freesort_distance]:\n ")
cat("Fixed effects (log odds): ", fixed_effects, "\n")
cat("Fixed effects (odds ratios): ", odds_ratios, "\n")
cat("Fixed effects (probabilities, causal_distance = 0):", probabilities, "\n")

```

## predicting effect sizes (i.e. proportion differences) based on differences in rating scores

### first get predicted score differences

```{r}
#getting predictions for causal condition (most causal item)
diffs_causal <-d_freesort_and_causal_distances %>%
  group_by(itemB, domain_itemA) %>%
  slice(which.max(euclidean_minus_causal)) %>% 
  ungroup() %>% 
  group_by(itemB) %>% 
  slice(which.max(euclidean_minus_causal)) %>% 
  ungroup() %>% 
  rename("causal_itemA" = itemA,
         "domain_causal_itemA" = domain_itemA,
         "causal_item_euc_minus_causal" = euclidean_minus_causal) %>%
  select(causal_itemA, itemB, domain_causal_itemA, domain_itemB, causal_item_euc_minus_causal)

#get predictions for causal condition (most causal item) 
diffs_similarity <- d_freesort_and_causal_distances %>%
  group_by(itemB, domain_itemA) %>%
  slice(which.min(euclidean_minus_causal)) %>% #least causal item??
  ungroup() %>% 
  mutate(same_domain = ifelse(domain_itemA == domain_itemB, TRUE, FALSE)) %>% 
  filter(same_domain == TRUE) %>% 
  ungroup() %>% 
  rename("similarity_itemA" = itemA,
       "domain_similarity_itemA" = domain_itemA,
       "similarity_item_euc_minus_causal" = euclidean_minus_causal) %>% 
  select(similarity_itemA, itemB, domain_similarity_itemA, domain_itemB, similarity_item_euc_minus_causal)

d_diffs <- diffs_causal %>% 
  left_join(diffs_similarity %>% 
              select(everything()), 
              by = c("itemB", "domain_itemB")) %>% 
  mutate(diff = causal_item_euc_minus_causal - similarity_item_euc_minus_causal)

```

### then plot predicted score differences

```{r}
dat_combined <- dat_summarized_by_item %>%
  group_by(condition, itemB) %>%
  reframe(
    prop_causal = mean(mean_proportion[response == "causal"], na.rm = TRUE),
    prop_similar = mean(mean_proportion[response == "similarity"], na.rm = TRUE),
    prop_causal = replace_na(prop_causal, 0),  #fill missing values with 0: (for some items all subj choose 0, causing NaNs or their proportions)
    prop_similar = replace_na(prop_similar, 0), 
    observed_diff = case_when(
      condition == "intervention" ~ prop_causal - prop_similar,  
      condition == "similarity" ~ prop_similar - prop_causal  
    )
  ) %>%
  left_join(d_diffs %>% select(itemB, diff), by = "itemB")  # merge in predicted differences

```

## correlation bar plot

```{r}
itemB_colors <- dat_summarized_by_item %>%
  distinct(itemB, domain_itemB) %>%
  mutate(color = case_when(
    domain_itemB == "bio" ~ "green",
    domain_itemB == "mind" ~ "blue",
    domain_itemB == "action" ~ "red"
  )) %>%
  pull(color, name = itemB)

barplot <- dat_combined %>%
  ggplot(aes(x = itemB)) +
  geom_bar(aes(y = observed_diff), stat = "identity", fill = "blue", alpha = 0.6) +
  geom_point(aes(y = diff), color = "red", size = 3) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = itemB_colors[levels(diffs_causal$itemB)])
  ) +
  labs(
    title = "Observed vs Predicted Proportion Differences by Item",
    x = "Item B",
    y = "observed proportion differences"
  ) +
  facet_wrap(~condition)

barplot

```


## correlation scatter plot 

```{r}
# Scatter plot with correlation for each condition
scatter_plot <- dat_combined %>%
  ggplot(aes(x = diff, y = observed_diff, color = condition)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed") +
  theme_bw() +
  facet_wrap(~condition) +
  labs(
    title = "Observed vs Predicted Differences by Condition",
    x = "Predicted proportion difference",
    y = "Observed proportion difference",
    color = "Condition"
  )

scatter_plot
```

### quantitative correlations 

```{r}
overall_correlation <- dat_combined %>%
  summarize(
    correlation = cor(diff, observed_diff, use = "complete.obs"),
    p_value = cor.test(diff, observed_diff, use = "complete.obs")$p.value,
    .groups = "drop"
  )

print(overall_correlation)

condition_level_correlations <- dat_combined %>%
  group_by(condition) %>%
  summarize(
    correlation = cor(diff, observed_diff, use = "complete.obs"),
    p_value = cor.test(diff, observed_diff, use = "complete.obs")$p.value,
    .groups = "drop"
  )

print(condition_level_correlations)

```


# Plotting logistic regression models

## Intervention condition

```{r}
#create a new dataset for prediction
prediction_data <- data.frame(
  causal_distance = seq(min(causal_condition$causal_distance), 
                        max(causal_condition$causal_distance), 
                        length.out = 100)
)

#add predicted probabilities to the dataset
prediction_data$predicted_prob <- predict(
  glm(formula = responses ~ causal_distance, family = binomial(link = "logit"), data = causal_condition),
  newdata = prediction_data,
  type = "response"
)

#plot 1: causal distance
ggplot(data = causal_condition, aes(x = causal_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving a Causal Response in Intervention Condition",
    y = "Probability",
    color = "Actual Response"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

#plot 2: freesort distance
ggplot(data = causal_condition, aes(x = freesort_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving a Causal Response in Intervention Condition",
    y = "Probability",
    color = "Actual Response"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

```

## similarity condition

```{r}
#plot 1: freesort distance
ggplot(data = similarity_condition, aes(x = freesort_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving Causal Response in Similarity Condition",
    y = "Prob",
    color = "observed response"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

#plot 2: causal distance
ggplot(data = similarity_condition, aes(x = causal_distance, y = responses)) +
  geom_point(aes(color = factor(responses)), alpha = 0.5, position = position_jitter(height = 0.02)) +
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE,  # Add confidence interval
    color = "blue",
    size = 1.2
  ) +
  labs(
    title = "Probability of Giving Causal Response in Similarity Condition",
    y = "Prob",
    color = "observed responses"
  ) +
  theme_bw() +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "top")

```












