---
title: "study 2"
output: html_document
date: "2024-10-28"
---

# ReadMe

This is the analysis for study 3 of the MBC project, alias the counterfactual study.  
The study had N = 50 (before exclusions), and subjects saw two conditions: Intervention and Similarity, with 15 trials per condition.
The question: 

# Set options

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(
               lme4, 
               here, 
               patchwork, 
               cowplot, 
               viridis, 
               effects, 
               gt,
               webshot2,
               knitr,
               performance,
               tidyverse)
sessionInfo()

```

# Global variables

```{r}
base_path <- here("code", "analysis", "pilot")
freesort_and_causal_distances_df_path <- paste0(base_path, "/data-explanation_1/freesort_minus_causals.csv")

```

# Study 3 - Counterfactuals

## Prepare data

### read data

```{r, warnings = FALSE}
data_path_cf <- paste0(base_path, "/data-counterfactual_1/pilot3")

desired_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "sit down", "jump up and down", "kick something", "take a walk", 
                   "get tired", "become hungry", "feel scared", "experience pain", "get sick")

file_paths <- list.files(path = data_path_cf, pattern = "^2025-", full.names = TRUE)

d <- imap_dfr(file_paths, ~ {
  read_csv(.x, col_types = cols(response = col_character())) %>%
    mutate(
      filename = basename(.x),
      subject_id = paste0("subj_", str_pad(.y, 2, pad = "0"))
    )}) %>%
  select(-c(mturk_participant_id, mturk_assignment_id, mturk_project_id)) %>%
  relocate(subject_id, .before = "success")

```

### wrangle data

initial: 30 trials * 48 subjects = 1440 rows.
dat: (30 trials / 2 trials per row) * 48 subjects = 720 rows.

TODO: rename itemB to target item.

```{r}
subjects_to_exclude <- d %>%
  filter(grepl("attention_check", condition)) %>% 
  filter(attn_correct == "no") %>% 
  pull(subject_id) %>% 
  unique()

dat <- d %>% 
  filter(!subject_id %in% subjects_to_exclude) %>%
  filter(condition %in% c("counterfactual", "similarity")) %>% 
  mutate(condition = ifelse(condition == "counterfactual", "causal", condition)) %>% #meant to call it causal (both are counterfactuals)
  mutate(response = as.factor(response)) %>% 
  mutate(condition = as.factor(condition)) %>% 
  select(-condition) %>% 
  select(subject_id, itemB, causal_itemA, similar_itemA, response_value) %>% 
  mutate(causal_itemA = factor(causal_itemA, levels = desired_order), #this col has unique values; we're assigning 15 lvls
         similar_itemA = factor(similar_itemA, levels = desired_order), #this col has unique values; we're assigning 15 lvls
         itemB = factor(itemB, levels = desired_order)) %>% 
  # filter(subject_id == "subj_01" & itemB == "see something") %>% 
  mutate(item_type = case_when(
        !is.na(causal_itemA) ~ "causal_itemA",
        !is.na(similar_itemA) ~ "similar_itemA"),
        item = coalesce(as.character(causal_itemA), as.character(similar_itemA))) %>% 
  select(-causal_itemA, -similar_itemA) %>% 
  pivot_wider(names_from = item_type,
              values_from = c(item, response_value)) %>% 
  rename("causal_itemA" = item_causal_itemA,
         "similar_itemA" = item_similar_itemA, 
         "causal_itemA_counterfactual" = response_value_causal_itemA,
         "similar_itemA_counterfactual" = response_value_similar_itemA) 

```

## plot dv vs conditions

### prepare to plot

```{r}
dat_plot <- dat %>% 
  rename("causal" = causal_itemA_counterfactual,
         "similar" = similar_itemA_counterfactual) %>% 
  pivot_longer(cols = c("causal", "similar"), 
               names_to = "counterfactual_type",
               values_to = "likelihood") 

```

### plot dv vs counterfactual type

```{r}
dat_plot %>% 
  ggplot(aes(x = counterfactual_type, y = likelihood)) +
  geom_violin() +
  geom_jitter(width = .05, alpha = .1) +
  stat_summary(fun = mean, geom = "point", size = 5, color = "red") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2) +
  geom_hline(yintercept = 50, linetype = "dashed") +
  labs(y = "Likelihood of Counterfactual", x = "Preceding Item type") +
  scale_x_discrete(labels = c("counterfactual" = "Causal", "similarity" = "Similar")) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))

```


### plot, facet by target items

first prune text labels

```{r, fig.width = 12, fig.height = 12}
dat_plot_long <- dat_plot %>%
  pivot_longer(cols = c("causal_itemA", "similar_itemA"),
               names_to = "item_type",
               values_to = "item") %>%
  group_by(subject_id, itemB) %>%
  slice(-c(2, 3))

label_data <- dat_plot_long %>%
  ungroup() %>% 
  distinct(itemB, counterfactual_type, item)

dat_plot_long %>%
  ggplot(aes(x = counterfactual_type, y = likelihood, color = counterfactual_type)) +
  geom_violin() +
  geom_jitter(width = .05, alpha = .1) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "red") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .2) +
  geom_hline(yintercept = 50, linetype = "dashed") +
  geom_text(data = label_data, aes(label = item), y = 95, vjust = 1, size = 3.5) +
  labs(y = "Likelihood of Occurring", x = "Counterfactual type") +
  scale_x_discrete(labels = c("counterfactual" = "Causal", "similarity" = "Similar")) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "none") +
  facet_wrap(~itemB)

```

### plot single subject

```{r, fig.width = 12, fig.height = 12}
dat_plot_long %>% 
  filter(subject_id == "subj_01") %>% 
  ggplot(aes(x = counterfactual_type, y = likelihood, color = counterfactual_type)) +
  geom_point() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  geom_text(aes(label = item), y = 95, vjust = 1, size = 3.5) +
  labs(y = "Likelihood of Occurring", x = "Counterfactual type") +
  theme_bw() +
  theme(
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        legend.position = "none") +
  facet_wrap(~itemB)

```

## plot dv vs distances

### prepare append distances 

```{r}
distances <- read.csv(freesort_and_causal_distances_df_path)

dat_distances <- dat_plot_long %>% 
  ungroup() %>% 
  rename("itemA" = item) %>% 
  left_join(distances %>% 
              select(itemB, itemA, mean_causal_distance, mean_freesort_distance, freesort_minus_causal),
            by = c("itemB", "itemA")) %>% 
  mutate(mean_causal_distance = round(mean_causal_distance, 2),
         mean_freesort_distance = round(mean_freesort_distance, 2),
         freesort_minus_causal = round(freesort_minus_causal, 2)) %>% 
  mutate(itemB = factor(itemB, levels = desired_order))

```

### plot difference scores

15 similar item - target item pairs, 15 causal item - target item pairs. 
The more causal (right of x -axis), the less likely the counterfactual event will occur

```{r, fig.width = 12}
#make separate data frame with only distances and difference scores?
#or, get the mean responses and an error bar for it?

#parcel out labels
label_distances <- dat_distances %>%
  distinct(itemB, counterfactual_type, itemA, freesort_minus_causal)
    

#plot
dat_distances %>% 
  ggplot(aes(x = freesort_minus_causal, y = likelihood, color = counterfactual_type)) +
  geom_point(alpha = .2) +
  geom_text(data = label_distances, aes(label = itemB), y = 95, vjust = .5, size = 3.5, angle = 80) +
  geom_text(data = label_distances, aes(label = itemA), y = 85, vjust = .5, size = 3.5, angle = 80) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .01, color = "black") +
  stat_summary(fun = "mean", geom = "point", shape = 1, size = 3, color = "black") +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 50, linetype = "dashed") +
  theme_bw() +
  labs(y = "Likelihood of Occurring", x = "Freesort minus Causal Distance") +
  theme_bw() +
  theme(
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12))
```

### itemwise difference scores

```{r, fig.width = 12}
dat_distances %>% 
  ggplot(aes(x = freesort_minus_causal, y = likelihood, color = counterfactual_type)) +
  geom_point(alpha = .2) +
  geom_text(data = label_distances, aes(label = itemA), y = 90, size = 3.5) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = .03, color = "black") +
  stat_summary(fun = "mean", geom = "point", shape = 1, size = 2, color = "black") +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 50, linetype = "dashed") +
  theme_bw() +
  labs(y = "Likelihood of Occurring", x = "Freesort minus Causal Distance") +
  theme_bw() +
  theme(
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12)) +
  facet_wrap(~itemB)


```


### plot individual distances

```{r, fig.width = 8}
dat_distances %>% 
  pivot_longer(cols = c("mean_causal_distance", "mean_freesort_distance"),
               names_to = "distance_type", 
               values_to = "distances") %>% 
  ggplot(aes(x = distances, y = likelihood, color = distance_type)) +
  geom_jitter(width = .01, alpha = .2) +
  geom_smooth(method = "lm") +
  geom_hline(yintercept = 50, linetype = "dashed") +
  theme_bw() +
  labs(y = "Likelihood of Occurring", x = "Distance") +
  theme_bw() +
  theme(
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12)) +
  facet_wrap(~item_type)

```



