---
title: "causation"
output: html_document
date: "2024-10-28"
---

This script does the following
1. Takes original data
2. Separates out freesort data and saves the csv (data_freesort)
   (does not save out causation data.. hmm why?)
3. Separates out combined (i.e. freesort + causation) data and saves the csv (data_pilot_12_14_combined)
4. Generates theoretical RDM data and saves (d_theories_categories)
5. Adds cosine embeddings and saves (d_all_theories) Q: why not just save 5 and do not save 4? what's diff btn 4 and 5?
6. Plots RDMs and saves them into /figures/

Changes made on 1/7/25
1. Moving data_freesort.csv and data_pilot_...csv over to ---> /snapshots_pre-processing_../
2. Moving theoretical rdm data from /code/analysis/pilot/theoretical-rdm-data/ ---> /snapshots_pre-processing_../
4. Rename final_output_csv to combined_output_csv?
5. Rename data_freesort to data_pilot_..._freesort

Study description
- Pilot ran on 12/13/24 on Connect CloudResearch, 50 participants
- Subject selection criteria: 95% approval rate, minimum 10 previous submissions, and fluent in English
- demographics: Age range

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
pacman::p_load(tidyverse, here, jsonlite)
sessionInfo()

library(jsonlite)
```

# Read data

```{r}
base_path <- here("code", "analysis", "pilot")
data_path <- paste0(base_path, "/data-combined/pilot05_121324_n50_combined_17")

file_paths <- list.files(path = data_path, pattern = "^2024-", full.names = TRUE)
d <- map_df(file_paths, ~ read_csv(.x) %>% mutate(filename = basename(.x))) %>% 
    filter(!str_detect(filename, "XX-XX-XX")) 

```

# parameters

```{r}
items <- c("choose what to do", "reach for something", "feel scared",  #items in exact order of img1.png to img15.png
           "experience pain", "get sick", "get tired", "become hungry", 
           "see something", "think about something", "hear something", 
           "remember something", "jump up and down", "kick something", 
           "take a walk", "sit down")

#exports
rdm_figures_outputs_path <- paste0(base_path, "/figures/theoretical-rdms/")

#freesort_output_path <- paste0(data_path, "/outputs/data_freeesort.csv")
#combined_output_csv <- "data_pilot_12_14_combined"
#combined_output_path <- paste0(data_path, "/outputs/", combined_output_csv, ".csv") 

combined_output_csv <- "pilot_121424_combined"
snapshots_path <- paste0(base_path, "/snapshots_preprocessing_analysis")

freesort_output_path <- paste0(snapshots_path, "/pilot_121424_freeesort.csv")
combined_output_path <- paste0(snapshots_path, "/", combined_output_csv, ".csv") 
theories_output_path <- paste0(snapshots_path, "/pilot_121424_theories.csv") #RENAMED FROM d_theories_categories
theories_all_output_path <- paste0(snapshots_path, "/pilot_121424_theories_all.csv") #RENAMED FROM d_all_theories

```


# PARSING

# freesort data

```{r}
# Process the final_locations column
data_freesort <- d %>%
  filter(trial_type == "free-sort") %>% # Filter for free-sort trials
  select(filename, trial_index, final_locations) %>% # Select relevant columns
  mutate(parsed_location = map(final_locations, ~ fromJSON(.))) %>% # Parse JSON
  unnest(parsed_location) %>% # Unnest the list column
  mutate(
    img_number = as.numeric(str_extract(src, "\\d+")), # Extract image numbers from `src`
    item = items[img_number] # Map image numbers to item labels
  ) %>%
  select(-c("final_locations", "src", "img_number")) %>% # Drop unnecessary columns
  mutate(trial_type = "free_sort") %>% # Add a new column for trial type
  relocate(trial_type, .after = filename) %>% # Place `trial_type` column next to `filename`
  rename(original_trial_index = trial_index)

write.csv(data_freesort, file = freesort_output_path, row.names = F)

```

# causation data

```{r}
response_data <- d %>%
  filter(trial_type == "survey-slider") %>% 
  select(filename, trial_index, response) %>%
  mutate(response = map(response, ~ tryCatch(
    fromJSON(.) %>%
      enframe(name = "question_id", value = "response"),
    error = function(e) tibble(question_id = character(), response = character())
  ))) %>%
  unnest(response) %>% # Unnest the response JSON
  mutate(
    response = unlist(response), # Unlist response values
    response = as.numeric(response), # Convert to numeric
    question_index = str_remove(question_id, "Q") # Remove "Q" prefix
  ) %>%
  select(filename, trial_index, question_index, response) # Include trial_indexs

metadata_data <- d %>%
  select(filename, trial_index, `0`:`14`, `15`) %>% # Include trial_index
  pivot_longer(
    cols = -c(filename, trial_index),  # Exclude filename and trial_index from long format
    names_to = "question_index",       # Convert column names to indices
    values_to = "metadata"             # Column for metadata JSON
  ) %>% 
  mutate(
    parsed_metadata = map(metadata, ~ {
      if (is.na(.x)) {
        tibble() # Handle NA metadata
      } else {
        fromJSON(.x, simplifyVector = FALSE) %>%
          map(~ if (is.null(.)) NA else .) %>%  # Replace NULL with NA
          as_tibble()
      }
    })
  ) %>% 
  unnest(parsed_metadata) %>%    # Unnest the parsed metadata columns
  select(-metadata)             # Drop the original metadata column for clarity

# Merge the response_data and metadata_data into data_causation
data_causation <- response_data %>%
  inner_join(metadata_data, by = c("filename", "trial_index", "question_index")) %>% 
  mutate(trial_type = "causation") %>% 
  relocate(trial_type, .after = filename) %>% # Place `trial_type` column next to `filename`
  rename(original_trial_index = trial_index, 
         original_question_index = question_index) %>% 
  group_by(filename, trial_type) %>%
  mutate(trial_index = dense_rank(original_trial_index),
         question_index = row_number()) %>%
  ungroup() %>% 
  relocate(c("trial_index", "question_index"), .after = "trial_type") %>% 
  relocate(c("original_trial_index", "original_question_index"), .after = "correct_response")

```

# Exclusions

- For causation we have 4 attention checks at pages 4, 8, 10 and 13. No attn checks for freesort 
- We are excluding subjects who fail at least 2 of them 
- Other posible criteria: time taken to complete study. (a subject took 13 minutes, which is maybe too short, i took ~20)

```{r}
data_causation <- data_causation %>%
  mutate(attention_check_id = case_when(
    attention_check == TRUE & trial_index == 4  ~ 1, 
    attention_check == TRUE & trial_index == 8  ~ 2, 
    attention_check == TRUE & trial_index == 10 ~ 3, 
    attention_check == TRUE & trial_index == 13 ~ 4, 
    TRUE ~ NA_integer_ #non-attention check rows
  )) %>%
  mutate(correct_response = case_when(
    attention_check == TRUE & trial_index == 4  ~ 0, #"Move slider all the way to left"
    attention_check == TRUE & trial_index == 8  ~ 75,  #"Move slider to 'Maybe yes'"
    attention_check == TRUE & trial_index == 10 ~ 100, #"Move slider all the way to right"
    attention_check == TRUE & trial_index == 13 ~ 0,   #"Move slider to 'Definitely not'"
    TRUE ~ correct_response # Retain existing values for non-attention check rows
  )) %>%
  mutate(correct = case_when(
    attention_check == TRUE & trial_index == 4  & response >= 0 & response <= 5   ~ TRUE, 
    attention_check == TRUE & trial_index == 8  & response >= 70 & response <= 80 ~ TRUE, 
    attention_check == TRUE & trial_index == 10 & response >= 0 & response <= 5 ~ TRUE, #3 and 4 are switched for some reason
    attention_check == TRUE & trial_index == 13 & response >= 95 & response <= 100   ~ TRUE,
    attention_check == TRUE ~ FALSE, #outside acceptable range
    TRUE ~ NA #non-attn check rows
  ))

excluded_participants <- data_causation %>%
  filter(attention_check == TRUE) %>% 
  group_by(filename) %>% 
  summarize(
    num_incorrect = sum(correct == FALSE, na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  filter(num_incorrect >= 2) %>% #identify participants with x or more incorrect (4 subjects had 1 incorrect attn check)
  pull(filename) 

data_causation <- data_causation %>%
  filter(!filename %in% excluded_participants)

```

Merging the two, adding condition order columns

# good!

```{r}
# Combine trial types to compare their starting original_trial_index
trial_order_info <- bind_rows(
  data_causation %>% distinct(filename, trial_type, original_trial_index),
  data_freesort %>% distinct(filename, trial_type, original_trial_index)
) %>%
  group_by(filename) %>%
  summarize(
    trial_type_order = case_when(
      min(original_trial_index[trial_type == "causation"]) < min(original_trial_index[trial_type == "free_sort"]) ~ 
        list(c(causation = "first", free_sort = "second")),
      TRUE ~ list(c(causation = "second", free_sort = "first"))
    )
  ) %>%
  unnest_wider(trial_type_order)

# Add trial_type_order to data_causation
data_causation <- data_causation %>%
  left_join(
    trial_order_info %>% select(filename, causation),
    by = "filename"
  ) %>%
  rename(trial_type_order = causation) %>%
  relocate(trial_type_order, .after = trial_type)

# Add trial_type_order to data_freesort
data_freesort <- data_freesort %>%
  left_join(
    trial_order_info %>% select(filename, free_sort),
    by = "filename"
  ) %>%
  rename(trial_type_order = free_sort) %>%
  relocate(trial_type_order, .after = trial_type)

```

# Merge both data frames

```{r}
# Merge causation and free_sort dataframes
data_combined <- bind_rows(
  data_causation,
  data_freesort
)

assign(combined_output_csv, data_combined)

write.csv(data_combined, combined_output_path, row.names = F)

```

## Generate theoretical RDM data 

```{r}
d_theories_categories <- expand.grid(itemA = desired_order, itemB = desired_order) %>% #expand grid
  filter(itemA != itemB) %>%  # Remove self-pairs
  
  #theory 1: (3 groups) mind, body, action
  mutate(MindBodyAction_itemA = case_when(itemA == "hear something" ~ "mind", #labels for first item of pair
                               itemA == "choose what to do" ~ "mind",
                               itemA == "remember something" ~ "mind",
                               itemA == "think about something" ~ "mind",
                               itemA == "see something" ~ "mind",
                               itemA == "reach for something" ~ "act",
                               itemA == "sit down" ~ "act", 
                               itemA == "jump up and down" ~ "act",
                               itemA == "kick something" ~ "act",
                               itemA == "take a walk" ~ "act",
                               itemA == "get tired" ~ "bio", 
                               itemA == "become hungry" ~ "bio",
                               itemA == "feel scared"~ "bio",
                               itemA == "experience pain"~ "bio",
                               itemA == "get sick"~ "bio"),
         MindBodyAction_itemB = case_when(itemB == "hear something" ~ "mind", #labels for second item of pair
                               itemB == "choose what to do" ~ "mind",
                               itemB == "remember something" ~ "mind",
                               itemB == "think about something" ~ "mind",
                               itemB == "see something" ~ "mind",
                               itemB == "reach for something" ~ "act",
                               itemB == "sit down" ~ "act", 
                               itemB == "jump up and down" ~ "act",
                               itemB == "kick something" ~ "act",
                               itemB == "take a walk" ~ "act",
                               itemB == "get tired" ~ "bio", 
                               itemB == "become hungry" ~ "bio",
                               itemB == "feel scared"~ "bio",
                               itemB == "experience pain"~ "bio",
                               itemB == "get sick"~ "bio")) %>% 
  
  #theory 2: (6 groups) perceptual, cognition, object directed, non-object directed, bodily tied to simulation, bodily not tied to stimulation
   mutate(PerCogOdActSpActBstimBnstim_itemA = case_when(itemA == "hear something" ~ "per",
                               itemA == "see something" ~ "per",
                               itemA == "choose what to do" ~ "cog",
                               itemA == "remember something" ~ "cog",
                               itemA == "think about something" ~ "cog",
                               itemA == "reach for something" ~ "odAct",
                               itemA == "sit down" ~ "SpAct", 
                               itemA == "jump up and down" ~ "SpAct",
                               itemA == "kick something" ~ "odAct",
                               itemA == "take a walk" ~ "SpAct",
                               itemA == "get tired" ~ "BnS", 
                               itemA == "become hungry" ~ "BnS",
                               itemA == "feel scared"~ "BS",
                               itemA == "experience pain"~ "BS",
                               itemA == "get sick"~ "BnS",),
         PerCogOdActSpActBstimBnstim_itemB = case_when(itemB == "hear something" ~ "per",
                               itemB == "see something" ~ "per",
                               itemB == "choose what to do" ~ "cog",
                               itemB == "remember something" ~ "cog",
                               itemB == "think about something" ~ "cog",
                               itemB == "reach for something" ~ "odAct",
                               itemB == "sit down" ~ "SpAct", 
                               itemB == "jump up and down" ~ "SpAct",
                               itemB == "kick something" ~ "odAct",
                               itemB == "take a walk" ~ "SpAct",
                               itemB == "get tired" ~ "BnS", 
                               itemB == "become hungry" ~ "BnS",
                               itemB == "feel scared"~ "BS",
                               itemB == "experience pain"~ "BS",
                               itemB == "get sick"~ "BnS")) %>% 

      #theory 3: (2 groups) physical and non-physical events:
      mutate(PhysicalEthereal_itemA = case_when(itemA == "hear something" ~ "ethereal",
                               itemA == "see something" ~ "ethereal",
                               itemA == "choose what to do" ~ "ethereal",
                               itemA == "remember something" ~ "ethereal",
                               itemA == "think about something" ~ "ethereal",
                               itemA == "reach for something" ~ "phys",
                               itemA == "sit down" ~ "phys", 
                               itemA == "jump up and down" ~ "phys",
                               itemA == "kick something" ~ "phys",
                               itemA == "take a walk" ~ "phys",
                               itemA == "get tired" ~ "phys", 
                               itemA == "become hungry" ~ "phys",
                               itemA == "feel scared"~ "phys",
                               itemA == "experience pain"~ "phys",
                               itemA == "get sick"~ "phys"),
             PhysicalEthereal_itemB = case_when(itemB == "hear something" ~ "ethereal",
                               itemB == "see something" ~ "ethereal",
                               itemB == "choose what to do" ~ "ethereal",
                               itemB == "remember something" ~ "ethereal",
                               itemB == "think about something" ~ "ethereal",
                               itemB == "reach for something" ~ "phys",
                               itemB == "sit down" ~ "phys", 
                               itemB == "jump up and down" ~ "phys",
                               itemB == "kick something" ~ "phys",
                               itemB == "take a walk" ~ "phys",
                               itemB == "get tired" ~ "phys", 
                              itemB == "become hungry" ~ "phys",
                               itemB == "feel scared"~ "phys",
                               itemB == "experience pain"~ "phys",
                               itemB == "get sick"~ "phys")) %>% 
  
  #scoring the items
  mutate("theor1_Mind-Body-Action" = ifelse(MindBodyAction_itemA == MindBodyAction_itemB, 0, 1),
         "theor2_PerCog-OdActSpAct-BodyStimBodynoStim" = ifelse(PerCogOdActSpActBstimBnstim_itemA == PerCogOdActSpActBstimBnstim_itemB, 0, 1),
         "theor3_Physical-Ethereal" = ifelse(PhysicalEthereal_itemA == PhysicalEthereal_itemB, 0, 1)) 

write.csv(d_theories_categories, theories_output_path, row.names = F)

```

## Compute cosine similarities from USE embeddings

```{r}
#load USE embeddings
embedding_df <- readRDS(here("code/analysis/pilot/embeddings/item_embeddings.rds")) #read embeddings

d_theories <- d_theories_categories %>% 
  select(-c("MindBodyAction_itemA", "MindBodyAction_itemB", 
            "PerCogOdActSpActBstimBnstim_itemA","PerCogOdActSpActBstimBnstim_itemB", 
            "PhysicalEthereal_itemA", "PhysicalEthereal_itemB"))

# Step 3: Merge embeddings back into the long data frame
# Add embeddings for itemA and itemB
d_theories_with_embeddings <- d_theories %>%
  left_join(embedding_df, by = c("itemA" = "item")) %>%
  rename(embeddingA = embedding) %>%
  left_join(embedding_df, by = c("itemB" = "item")) %>%
  rename(embeddingB = embedding)

# Compute cosine similarity for each row
d_all_theories <- d_theories_with_embeddings %>%
  rowwise() %>% 
  mutate(cosine_similarity = sum(unlist(embeddingA) * unlist(embeddingB)) /
                 (sqrt(sum(unlist(embeddingA)^2)) * sqrt(sum(unlist(embeddingB)^2)))) %>%
  ungroup() %>% 
  rowwise() %>% 
  mutate(cosine_similarity = (1 - cosine_similarity) / 2) %>% #turning cosine similarity to a disimilarity measure. For some reason high cosine similarity means disimilar
  ungroup() %>% 
  select(-c("embeddingA", "embeddingB")) %>% 
  rename("theor4_cosine_similarity" = cosine_similarity) %>% 
  ungroup() 

# d_all_theories %>% 
#   arrange(theor4_cosine_similarity)
  
write.csv(d_all_theories, theories_all_output_path, row.names = F)

```

# plot theoretical RDMS

```{r}
plot_rdm_theory <- function(data, column_name, desired_order, plot_title) {
  # Prepare data and reorder
  df_matrix <- data %>%
    mutate(
      itemA = factor(itemA, levels = desired_order),
      itemB = factor(itemB, levels = desired_order)
    ) %>%
    select(itemA, itemB, all_of(column_name)) %>%
    pivot_wider(names_from = itemB, values_from = all_of(column_name)) %>%
    column_to_rownames("itemA")
  
  # Reorder rows and columns
  df_matrix <- df_matrix[desired_order, desired_order]
  diag(df_matrix) <- 0 # Set diagonal to 0
  
  # Plot RDM
  corrplot(
    as.matrix(df_matrix), method = "color", is.corr = FALSE,
    tl.col = "black", title = plot_title,
    mar = c(0, 0, 2, 0), addgrid.col = "darkgray"
  )
}

# Define desired orders
mba_order <- c("see something", "hear something", "choose what to do", 
               "remember something", "think about something", "reach for something", 
               "sit down", "jump up and down", "kick something", "take a walk", 
               "get tired", "become hungry", "feel scared", "experience pain", "get sick")

perccog_order <- c("see something", "hear something", "choose what to do", 
                   "remember something", "think about something", "reach for something", 
                   "kick something", "sit down", "jump up and down",  "take a walk", 
                   "get tired", "become hungry", "get sick", "feel scared", "experience pain")

physethereal_order <- c("see something", "hear something", "choose what to do", 
                        "remember something", "think about something", "reach for something", 
                        "kick something", "sit down", "jump up and down",  "take a walk", 
                        "get tired", "become hungry", "get sick", "feel scared", "experience pain")

# Plot RDMs for each theory
plot_rdm_theory(d_all_theories, "theor4_cosine_similarity", mba_order, "1. Cosine Distance - RDM")
plot_rdm_theory(d_all_theories, "theor1_Mind-Body-Action", mba_order, "2. Mind, Body & Action - RDM")
plot_rdm_theory(d_all_theories, "theor2_PerCog-OdActSpAct-BodyStimBodynoStim", perccog_order, "3. Perception, Cognitive et al. - RDM")
plot_rdm_theory(d_all_theories, "theor3_Physical-Ethereal", physethereal_order, "4. Physical & Ethereal - RDM")


# save them
#output_folder <- paste0(here(rdm_outputs_path), "/figures/theoretical-rdms/") -> delete this

theories <- list(
  list(column_name = "theor4_cosine_similarity", order = mba_order, title = "1. Cosine Distance - RDM"),
  list(column_name = "theor1_Mind-Body-Action", order = mba_order, title = "2. Mind, Body & Action - RDM"),
  list(column_name = "theor2_PerCog-OdActSpAct-BodyStimBodynoStim", order = perccog_order, title = "3. Perception, Cognitive et al. - RDM"),
  list(column_name = "theor3_Physical-Ethereal", order = physethereal_order, title = "4. Physical & Ethereal - RDM")
)

for (i in seq_along(theories)) {
  theory <- theories[[i]]
  output_file <- paste0(rdm_figures_outputs_path, "theory_", i, ".pdf")
  pdf(file = output_file, width = 8, height = 6)   # Open the PDF device
  plot_rdm_theory(
    data = d_all_theories,
    column_name = theory$column_name,
    desired_order = theory$order,
    plot_title = theory$title
  )
  dev.off() # Close the PDF device
}

```

Demographics

```{r}
# demographics
df_demographics <- d %>% 
  filter(trial_id == "demographics_survey") %>% 
  select("mturk_participant_id", "trial_id", "filename", 
         "feedback", "age", "ethnicity", "gender", "race") 

summary(df_demographics$age)
table(df_demographics$gender)
table(df_demographics$ethnicity)
table(df_demographics$race)

```




Scrap on attention check

```{r}
d %>% 
  distinct(trial_id)
  filter(trial_id )

```



